---
title: "Practical 2: R Exercises II -- Solutions"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
---

> Try the exercises yourself first. The value is in getting stuck, not in reading the answer.


## Part 1: Writing your own functions

### Exercise 1

```{r}
celsius_to_fahrenheit <- function(c) {
  f <- c * 9/5 + 32
  return(f)
}

celsius_to_fahrenheit(0)
celsius_to_fahrenheit(37)
celsius_to_fahrenheit(100)
```

### Exercise 2

```{r}
mean_center <- function(x) {
  x - mean(x)
}

mean_center(c(452, 618, 390, 705, 511))
```

### Exercise 3

```{r}
standardize <- function(x, target_mean = 0, target_sd = 1) {
  centered <- mean_center(x)
  z <- centered / sd(x)
  z * target_sd + target_mean
}

standardize(c(452, 618, 390, 705, 511))          # default: mean 0, SD 1
standardize(c(452, 618, 390, 705, 511), 100, 15)  # rescaled to mean 100, SD 15
```

------------------------

## Part 2: If-statements

### Exercise 4

```{r}
classify_score <- function(score) {
  if (score >= 55) {
    "pass"
  } else {
    "fail"
  }
}

classify_score(40)
classify_score(55)
classify_score(80)
```

### Exercise 5

```{r}
classify_grade <- function(score) {
  if (score >= 85) {
    "A"
  } else if (score >= 70) {
    "B"
  } else if (score >= 55) {
    "C"
  } else {
    "F"
  }
}

classify_grade(90)
classify_grade(72)
classify_grade(60)
classify_grade(30)
```

### Exercise 6 (discussion)

```{r}
# classify_score(c(40, 55, 80))   # try running this -- it errors (or in older R, warns
                                    # and silently uses only the first element)
```

`if` expects a single `TRUE`/`FALSE` value, not a whole vector of them
-- when you hand it a vector of three comparisons, it doesn't know which
one to branch on, so it errors (in current R versions) rather than
guessing. `ifelse()` is specifically built to handle a whole vector at
once: it checks each element's condition individually and returns a
vector of results, one per input. The rule of thumb: use `if`/`else` for
a single decision inside your code's logic (like deciding what a
function should do); use `ifelse()` when you want to apply the same
decision to every element of a vector at once, as a vectorized
operation.

--------------------

## Part 3: For-loops

### Exercise 7

```{r}
for (i in 1:10) {
  print(i^2)
}
```

### Exercise 8

```{r}
scores <- c(45, 78, 62, 90, 38, 71)

running_total <- 0
for (i in 1:length(scores)) {
  running_total <- running_total + scores[i]
}
running_total
```

You should get `384`, matching `sum(scores)`.

### Exercise 9

```{r}
count_above_60 <- 0
for (i in 1:length(scores)) {
  if (scores[i] > 60) {
    count_above_60 <- count_above_60 + 1
  }
}
count_above_60
```

You should get `4`.

---------------------------------

## Part 4: Matrices

### Exercise 10

R fills column by column, so `matrix(c(1, 3, 2, 4), nrow = 2)` places
the numbers `1, 3` down the first column and `2, 4` down the second:

```{r}
M <- matrix(c(1, 3, 2, 4), nrow = 2)
M
```

Row 1, column 2 is `2`.

### Exercise 11

```{r}
M2 <- matrix(1:9, nrow = 3)
M2

M2[2, 3]      # (a) row 2, column 3
M2[2, ]       # (b) entire second row
M2[, 3]       # (c) entire third column
```

### Exercise 12

```{r}
W <- matrix(c(1, 3, 2, 4), nrow = 2)
x <- c(5, 6)

W * x     # element-wise: [1*5, 3*6, 2*5, 4*6] reshaped -> completely different numbers
W %*% x   # matrix multiplication: row 1 . x = 17, row 2 . x = 39
```

`W %*% x` gives `17` and `39`, matching the worked hand-calculation in
the practical exactly: row 1 of `W` (`1, 2`) combined with `x` (`5, 6`)
is `1*5 + 2*6 = 17`; row 2 (`3, 4`) combined with `x` is
`3*5 + 4*6 = 39`. `W * x` gives something else entirely, because it's
not pairing rows with columns at all -- it's just multiplying whatever
happens to sit in the same position once `x` is recycled down `W`'s
column-major storage order. Worth actually printing both outputs side by
side and confirming they don't match -- this is the single most common
source of silently-wrong results once matrices show up for real in the
neural network sessions.


--------------------------------------

## Part 5: For-loops and vector arithmetic

### Exercise 13

```{r}
observed  <- c(0.40, 0.60, 0.55, 0.60, 0.62, 0.90, 0.92, 0.95, 1.00)
predicted <- c(0.50, 0.60, 0.70, 0.75, 0.80, 0.85, 0.88, 0.90, 0.93)

total_squared_error <- 0
for (i in 1:length(observed)) {
  squared_error <- (observed[i] - predicted[i])^2
  total_squared_error <- total_squared_error + squared_error
}
rmsd_loop <- sqrt(total_squared_error / length(observed))

rmsd_vectorized <- sqrt(mean((observed - predicted)^2))

rmsd_loop == rmsd_vectorized
```

You should get `TRUE` -- both approaches compute exactly the same
number.

### Exercise 14 (discussion)

`(observed - predicted)` subtracts the two vectors element by element,
giving a vector of 9 differences (one per data point) -- exactly what
the loop's `observed[i] - predicted[i]` computes, one `i` at a time,
except all 9 at once. `^2` then squares every one of those 9 differences
simultaneously, matching the loop's `squared_error <- (...)^2` line
applied to all elements together rather than one per iteration. `mean()`
then does two things the loop needed two separate pieces of code for: it
adds all 9 squared differences together (replacing the loop's
`total_squared_error <- total_squared_error + squared_error` running
total), *and* divides by how many there are (replacing the loop's final
`/ length(observed)`) -- both folded into one function call.

### Exercise 15

```{r}
mean_center_loop <- function(x) {
  m <- mean(x)
  result <- numeric(length(x))
  for (i in 1:length(x)) {
    result[i] <- x[i] - m
  }
  result
}

mean_center_loop(c(452, 618, 390, 705, 511))
mean_center(c(452, 618, 390, 705, 511))
```

Both should give identical output.

### Exercise 16 (discussion)

The vectorized version is easier to read at a glance --
`sqrt(mean((observed - predicted)^2))` states the RMSD formula almost
exactly as you'd write it in mathematical notation, in one line, while
the loop version takes 5 lines to express the same idea and requires
holding "what's `total_squared_error` currently equal to" in your head
as you read through it. The loop version also has more places for a
small bug to hide: starting the running total at the wrong value, an
off-by-one error in `1:length(observed)`, forgetting to reset the total
before a second run -- none of these are possible in the vectorized
version, because there's no running state to manage at all. Speed
matters too, especially once datasets get large, but the readability and
bug-avoidance reasons alone are why essentially every function you'll
write for the rest of this course -- `rmsd()`, `nll()`,
`psychometric_model()` -- is written the vectorized way, with no
explicit loop over data points anywhere in sight.
