---
title: "Practical 2: Introduction 2"
subtitle: "Functions, Control Flow, Matrices, and Vectorization"
output:
  html_document:
    toc: true
---

## Before you start

Practical 1 covered *calling* R's built-in functions and working with
vectors. This session covers five things that don't show up anywhere in
our manual, but that you'll use constantly for the rest of this
course: **writing your own functions**, **if-statements**,
**for-loops**, **matrices**, and — tying everything together — **why a
single line of vector arithmetic can replace an entire loop**. By the
end, you'll be able to read (and soon write) code that looks exactly
like what you'll see in every practical from here on.

-------------------

## Part 1: Writing your own functions

You've called plenty of built-in functions already (`sqrt()`, `mean()`,
`round()`...). Now you'll write your own. The basic shape:

``` r
my_function <- function(argument1, argument2) {
  result <- argument1 + argument2   # do something with the arguments
  return(result)                     # send the result back out
}
```

A function is just a named, reusable recipe: give it some inputs (its
*arguments*), it does something with them, and it *returns* a value.
`return()` is technically optional — a function returns whatever its
last line evaluates to — but writing it explicitly makes it obvious
what's being sent back, especially once your functions get longer.

### Exercise 1

Write a function `celsius_to_fahrenheit(c)` that
converts a Celsius temperature to Fahrenheit ($F = C \times 9/5 + 32$).
Test it on 0, 37 (body temperature), and 100.

```{r}
# Your code here

```

### Exercise 2

Write a function `mean_center(x)` that takes a numeric
vector and returns it with its own mean subtracted off (exactly what you
did by hand in Practical 1, Exercise 11 — now wrapped in a reusable
function).

```{r}
# Your code here

```

### Exercise 3

Functions can have **default values** for their
arguments, so you don't have to specify every argument every time
(you've already used this — recall `rnorm(n, mean = 0, sd = 1)` from
Practical 1). Write `standardize(x, target_mean = 0, target_sd = 1)`
that rescales a vector to have a chosen mean and SD (default: mean 0, SD
1 — a "z-score"). *(Hint: first mean-center `x`, then divide by its SD,
then multiply by `target_sd` and add `target_mean`.)*

```{r}
# Your code here

```

-------------------

## Part 2: If-statements

An if-statement runs a block of code only when a condition is `TRUE`:

``` r
if (condition) {
  # runs only if condition is TRUE
} else {
  # runs only if condition is FALSE
}
```

The `else` part is optional — you can have an `if` with no `else` at
all.

### Exercise 4
Write a function `classify_score(score)` that returns
`"pass"` if `score >= 55` and `"fail"` otherwise. Test it on 40, 55, and
80.

```{r}
# Your code here

```

### Exercise 5

Extends Exercise 4 into `classify_grade(score)`, using
**chained** `if`/`else if`/`else` to return `"A"` for 85+, `"B"` for
70–84, `"C"` for 55–69, and `"F"` below 55.

```{r}
# Your code here

```

**Discussion** Exercise 6 from Practical 1 used `ifelse()`
to avoid writing a full if-statement. Now that you know both: what's
actually different between `ifelse()` and `if`/`else`? Try running
`classify_score()` (your Exercise 4 function, using `if`/`else`) on the
whole vector `c(40, 55, 80)` at once, instead of one number at a time.
What happens, and why do you think that is?

-------------------

## Part 3: For-loops

A for-loop repeats a block of code once for every element of a sequence:

``` r
for (i in 1:5) {
  print(i)
}
```

`i` takes each value in `1:5` in turn — first 1, then 2, and so on —
running the code inside `{}` each time.

### Exercise 7

Write a for-loop that prints the square of every number
from 1 to 10.

```{r}
# Your code here

```

### Exercise 8

Given `scores <- c(45, 78, 62, 90, 38, 71)`, use a
for-loop (not `sum()`) to add up all the values, one at a time, into a
running total.

```{r}
scores <- c(45, 78, 62, 90, 38, 71)
# Your code here

```

### Exercise 9

Combine a for-loop with an if-statement: count how many
values in `scores` (Exercise 8) are above 60, by looping through each
one and incrementing a counter when the condition is met.

```{r}
# Your code here

```

-------------------

## Part 4: Matrices

A matrix is a vector with rows *and* columns — our manual mentions
them in one sentence and moves on, but you'll need them constantly from
here on (every neural network practical later in this course represents
a whole layer's weights as a matrix).

```{r}
M <- matrix(c(1, 3, 2, 4), nrow = 2)
M
```

### Exercise 10

Look at the output above. R fills a matrix **column by
column**, not row by row. Given `matrix(c(1, 3, 2, 4), nrow = 2)`, work
out by hand what each element's position is *before* running the code —
which value ends up in row 1, column 2? Check your answer against the
output.

### Exercise 11

Build a 3×3 matrix containing the numbers 1 through 9,
filled column-by-column. Then use `[row, column]` indexing to extract:
(a) the element in row 2, column 3; (b) the entire second row; (c) the
entire third column.

```{r}
# Your code here

```

### Excursus: What is matrix multiplication, actually?

If matrix algebra feels rusty, this is worth slowing down for —
everything from here to the end of the course leans on it.

Matrix multiplication combines a **row** from the first matrix with a
**column** from the second: multiply each pair of corresponding entries,
then add all those products together into a single number. That's it —
one number in, for one row paired with one column. Do this for *every*
row-column pairing, and you get the full result.

Take a small, concrete example:

$$W = \begin{pmatrix} 1 & 2 \\ 3 & 4 \end{pmatrix}, \qquad x = \begin{pmatrix} 5 \\ 6 \end{pmatrix}$$

$$W x = \begin{pmatrix} (1 \times 5) + (2 \times 6) \\ (3 \times 5) + (4 \times 6) \end{pmatrix}
      = \begin{pmatrix} 5 + 12 \\ 15 + 24 \end{pmatrix}
      = \begin{pmatrix} 17 \\ 39 \end{pmatrix}$$

Row 1 of $W$ (`1, 2`) combined with the column $x$ (`5, 6`) gives `17`.
Row 2 of $W$ (`3, 4`) combined with the *same* column gives `39`. Every
row of the first matrix gets combined with the (single) column of the
second, one at a time, each producing one number in the result.

**The dimension rule.** For this "multiply-then-add" pairing to make
sense, a row from the first matrix and a column from the second need to
be the **same length** — otherwise there's nothing to pair up. That's
the whole rule for whether two matrices can be multiplied at all: an
$(m \times n)$ matrix can be multiplied by an $(n \times p)$ matrix only
because the *inner* dimensions ($n$ and $n$) match — the result comes
out $(m \times p)$. Here, $W$ is $2\times2$ and $x$ is $2\times1$ (a
vector is just a matrix with one column) — inner dimensions both 2,
result is $2\times1$.

**This is genuinely different from element-wise multiplication (`*`).**
`*` doesn't pair a row with a column at all — it just multiplies
whatever's in the same position in both objects, one position at a time,
with no summing and no combining across positions. It needs the two
objects to already be the same shape (or for R to recycle a shorter one
to fit); matrix multiplication (`%*%`) instead needs the specific "row
length equals column length" relationship above, and produces a
genuinely new shape by combining values *across* positions, not just
matching them up.

Exercise 12 asks you to see this contrast directly, with real numbers.

### Exercise 12

This is the single most important gotcha with matrices
in R. Run both of these and compare the results:

```{r}
W <- matrix(c(1, 3, 2, 4), nrow = 2)
x <- c(5, 6)

W * x     # element-wise multiplication
W %*% x   # matrix multiplication
```

They give *completely* different answers. `W %*% x` gives `17` and `39`,
exactly matching the worked example above — each row of `W`, combined
with all of `x`. `W * x` instead multiplies matching positions one at a
time (recycling `x` down the matrix's columns, which can be genuinely
confusing to read, since it's silently using R's column-major storage
order underneath). **Whenever you actually mean matrix multiplication —
which is almost always, once you get to the neural network sessions,
where each row of a weight matrix is one hidden unit's own weights — you
need `%*%`, never `*`.** Getting this wrong doesn't usually throw an
error; it just silently gives you the wrong numbers, which is exactly
what makes it dangerous.

-------------------

## Part 5: For-loops and vector arithmetic are doing the same thing

Here's the payoff for everything above. These two chunks compute the
*exact same result* — the RMSD between some observed and predicted
values, a calculation you've already seen in the course and will keep
seeing:

```{r}
observed  <- c(0.40, 0.60, 0.55, 0.60, 0.62, 0.90, 0.92, 0.95, 1.00)
predicted <- c(0.50, 0.60, 0.70, 0.75, 0.80, 0.85, 0.88, 0.90, 0.93)

# Version 1: an explicit for-loop
total_squared_error <- 0
for (i in 1:length(observed)) {
  squared_error <- (observed[i] - predicted[i])^2
  total_squared_error <- total_squared_error + squared_error
}
rmsd_loop <- sqrt(total_squared_error / length(observed))
rmsd_loop
```

```{r}
# Version 2: vectorized -- no loop at all
rmsd_vectorized <- sqrt(mean((observed - predicted)^2))
rmsd_vectorized
```

### Exercise 13

Confirm `rmsd_loop` and `rmsd_vectorized` giveidentical results.

```{r}
# Your code here

```

### Exercise 14

**Discussion**: Walk through Version 1 line by line, and
explain in your own words what each piece of Version 2 is doing instead.
Specifically: what does `(observed - predicted)` alone give you (before
squaring)? What is `^2` doing to every element of that at once? What is
`mean()` doing that replaces both the for-loop's running total *and* the
final division by `length(observed)`?

### Exercise 15

Write a for-loop version of `mean_center()` from
Exercise 2 (looping over each element, subtracting the mean one at a
time into a new result vector), and confirm it gives the same output as
your original vectorized version.

```{r}
# Your code here

```

### Exercise 16

**Discussion**: If both versions always give the same
answer, why does this whole course — and essentially all real R code —
prefer the vectorized version whenever possible? Consider: which version
was easier to *read* at a glance? Which version would you trust more to
be free of small bugs (an off-by-one error in the loop bounds,
forgetting to reset the running total)? *(There's also a real speed
difference, which matters a lot for very large datasets — but even on
data too small to notice that, the readability and bug-avoidance reasons
alone are why you'll see almost no explicit for-loops once you get
further into this course's model-fitting code.)*
