---
title: "Practical 5: Neural Networks 2"
subtitle: "How Many Hidden Units Do You Actually Need?"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

## Before you start

In the lecture we built a 2-input, 2-hidden-unit network by hand,
trained it with backpropagation, and watched it solve XOR. The
`forward()` function you saw there already used matrix notation
(`W1 %*% x`), which means it was general the whole time, i.e., it never
actually assumed exactly 2 inputs or exactly 2 hidden units, those
numbers just came from the *shape* of `W1`.

Most of the code below is already written for you, including a fully
general training function. Your job in most exercises is to **run it,
then change something** (how many hidden units, which dataset, how many
restarts) and observe what happens. This practical is about building
intuition for *when* and *why* a network needs more hidden units.

You don't need to understand every line of R syntax perfectly, rather
focus on understanding *what each step is doing conceptually*. If
something errors, re-check you've run all the chunks above it in order.

## Part 0: A harder problem — 3-bit parity

XOR is "is exactly one of these two bits on?" The natural 3-input
generalization is **parity**: is the *number* of 1s among the three
inputs odd?

```{r}
# Every combination of 3 binary inputs (2^3 = 8 rows), same idea as the lecture's xor_data
# but with one more input column.
parity_data <- expand.grid(x1 = c(0, 1), x2 = c(0, 1), x3 = c(0, 1))
parity_data$category <- (parity_data$x1 + parity_data$x2 + parity_data$x3) %% 2  # 1 if the sum is odd
parity_data
```

### Exercise 1

Confirm by eye that `category` is 1 exactly when an odd number of `x1`,
`x2`, `x3` are 1 (1 or 3 of them), and 0 otherwise. This is a genuine
three-input generalization of last lecture's XOR.

## Part 1: A fully general training function

`sigmoid()` and `forward()` are unchanged from the lecture .

```{r}
sigmoid <- function(x) {
  1 / (1 + exp(-x))
}

forward <- function(x, W1, b1, w2, b2) {
  h <- sigmoid(W1 %*% x + b1)
  output <- sigmoid(sum(w2 * h) + b2)
  list(h = as.vector(h), output = output)
}
```

### Exercise 2

Sanity-check `forward()` with a single hidden unit to see if it still
works with one hidden unit instead of two:

```{r}
forward(x = c(1, 1), W1 = matrix(c(1, 1), nrow = 1), b1 = 0, w2 = 1, b2 = 0)
```

You should get `h` around 0.881 and `output` around 0.707. Now try
changing `w2` from `1` to `-1` and re-run. Does `output` go up or down,
and does that match what you'd expect from a negative output weight?

Here's a fully general training function, which generalizes the
lecture's `train_xor()`. It takes the input matrix `X` (it works out the
number of inputs itself, as `ncol(X)`) and `n_hidden` as arguments,
instead of assuming there are always 2 inputs and 2 hidden units:

```{r}
train_network <- function(X, y, n_hidden, seed, lr = 0.5, n_epochs = 2000) {
  set.seed(seed)
  X <- as.matrix(X)
  n_inputs <- ncol(X)

  # random starting weights, sized from n_hidden and n_inputs -- not hardcoded
  W1 <- matrix(rnorm(n_hidden * n_inputs, sd = 1), nrow = n_hidden)
  b1 <- rep(0, n_hidden)
  w2 <- rnorm(n_hidden, sd = 1)
  b2 <- 0

  loss_history <- numeric(n_epochs)
  for (epoch in 1:n_epochs) {
    for (i in 1:nrow(X)) {
      x <- X[i, ]; target <- y[i]
      fwd <- forward(x, W1, b1, w2, b2)
      h <- fwd$h; output <- fwd$output

      delta_out <- (target - output) * output * (1 - output)      # same delta rule as the lecture
      delta_hidden <- delta_out * w2 * h * (1 - h)                 # same backward-blame formula

      # outer(delta_hidden, x) updates every entry of W1 in one line -- it's the
      # same by-hand update from the lecture's worked example, just generalized to
      # however many hidden units there are, instead of writing out 4 fixed lines:
      #   for each hidden unit i, for each input j:
      #     W1[i, j] <- W1[i, j] + lr * delta_hidden[i] * x[j]
      W1 <- W1 + lr * outer(delta_hidden, x)
      b1 <- b1 + lr * delta_hidden
      w2 <- w2 + lr * delta_out * h
      b2 <- b2 + lr * delta_out
    }
    predictions <- numeric(nrow(X))                      # one prediction per row of the data
    for (i in 1:nrow(X)) {
      predictions[i] <- forward(X[i, ], W1, b1, w2, b2)$output
    }
    loss_history[epoch] <- sqrt(mean((y - predictions)^2))
  }
  list(W1 = W1, b1 = b1, w2 = w2, b2 = b2, loss_history = loss_history)
}
```

### Exercise 3

Try `train_network()` with only **one** hidden unit on the lecture's XOR
data:

```{r}
x1 <- c(0, 0, 1, 1); x2 <- c(0, 1, 0, 1); category <- c(0, 1, 1, 0)
xor_data <- data.frame(x1, x2, category)

fit_1hidden <- train_network(X = xor_data[, c("x1","x2")], y = xor_data$category,
                              n_hidden = 1, seed = 1)
fit_1hidden$loss_history[2000]
```

Does a single hidden unit manage to solve XOR? Should it, given what you
already know from last lecture about what one unit, whether it is hidden
or not, can and can't do? Keep this result in mind for Part 4.

## Part 2: Sanity-check on familiar ground

Before trusting `train_network()` on a new problem, check it still gets
the *old* problem right. As in the lecture, we train several networks from
different random starting points (restarts) and then pick one that
worked.

### Exercise 4

```{r}
xor_runs <- list()
xor_final_losses <- numeric(10)
for (seed in 1:10) {
  xor_runs[[seed]] <- train_network(xor_data[, c("x1","x2")], xor_data$category, n_hidden = 2, seed = seed)
  xor_final_losses[seed] <- xor_runs[[seed]]$loss_history[2000]
}
round(xor_final_losses, 4)

chosen_xor <- xor_runs[[9]]          # the run from seed 9
chosen_xor$loss_history[2000]
```

Several runs end well under 0.1, like in the lecture, and some are
stuck at a much higher RMSD. We'll look at the run from **seed 9**
(RMSD about 0.05, nearly as good as the best one), because its two
lines slope downward like the lecture's OR and NAND lines. (Exercise 4b
shows that other successful runs find a different, equally good
solution.) Now plot each hidden unit's own decision boundary, which is the
same formula as the lecture's "It works!" slide, row $i$ of `W1` gives
$x_2 = -(w_{i1}x_1 + b_i)/w_{i2}$:

```{r}
plot(xor_data$x1, xor_data$x2, pch = 19, cex = 3,
     col = ifelse(xor_data$category == 1, "steelblue", "darkorange"),
     xlab = "x1", ylab = "x2", xlim = c(-0.5, 1.5), ylim = c(-0.5, 1.5))
x_seq <- seq(-0.5, 1.5, length.out = 100)
for (i in 1:2) {
  y_line <- -(chosen_xor$W1[i, 1] * x_seq + chosen_xor$b1[i]) / chosen_xor$W1[i, 2]
  lines(x_seq, y_line, col = i + 1, lwd = 2)
}
legend("topright", legend = c("hidden unit 1", "hidden unit 2"), col = c(2, 3), lwd = 2)
```

You should see two lines that, combined, separate the XOR categories.
The same "two lines" picture from the lecture, just found automatically
this time instead of hand-picked.

Now look at how the error changed *during* training. `train_network()`
stores the RMSD after every epoch in `loss_history`, so we can plot the
learning curve of all 10 runs on one plot:

```{r}
plot(xor_runs[[1]]$loss_history, type = "l", lwd = 2, col = 1, ylim = c(0, 0.6),
     xlab = "epoch", ylab = "RMSD")
for (seed in 2:10) {
  lines(xor_runs[[seed]]$loss_history, lwd = 2, col = seed)
}
```

Describe what you see. Do all runs start at about the same error? Do
they all come down at the same speed? Do some sit on a flat stretch
(a plateau) for a long time before dropping? Are there runs that never
come down within 2000 epochs, and what does that tell you about what
"stuck" means?

### Exercise 4b: does every run find the same two lines?

The plot above shows the lines from just *one* run (seed 9). But
you trained 10 networks, each from a different random start. Plot the
two lines each run found, side by side:

```{r, fig.width = 12, fig.height = 5}
par(mfrow = c(2, 5))
for (r in 1:10) {
  fit <- xor_runs[[r]]
  plot(xor_data$x1, xor_data$x2, pch = 19, cex = 2,
       col = ifelse(xor_data$category == 1, "steelblue", "darkorange"),
       xlab = "x1", ylab = "x2", xlim = c(-0.5, 1.5), ylim = c(-0.5, 1.5),
       main = paste("seed", r, "- RMSD", round(xor_final_losses[r], 3)))
  for (i in 1:2) {
    y_line <- -(fit$W1[i, 1] * x_seq + fit$b1[i]) / fit$W1[i, 2]
    lines(x_seq, y_line, col = i + 1, lwd = 2)
  }
}
par(mfrow = c(1, 1))
```

And print each hidden unit's **slope** (from the line's equation above,
the slope is $-w_{i1}/w_{i2}$):

```{r}
for (r in 1:10) {
  fit <- xor_runs[[r]]
  slopes <- numeric(2)
  for (i in 1:2) {
    slopes[i] <- -fit$W1[i, 1] / fit$W1[i, 2]
  }
  cat("seed", r, " RMSD", round(xor_final_losses[r], 3), " slopes:", round(slopes, 2), "\n")
}
```

Ignore any run that got stuck (RMSD well above 0.1) as its lines don't
solve XOR. Among the runs that *did* solve it: do all of them find the
same two lines? Is each line sloping upward (positive slope) or downward
(negative slope)? Within a single run, do the two lines slope the same
way or opposite ways?

Then think about *why* both options can work. Sketch (or picture) the
four XOR points: which pairs of parallel lines could each trap the two
blue points between them, or push them to opposite outsides? What does
this say about whether a trained network's solution is *unique*?

**Optional:** train 50 networks instead of 10 and count how many
successful runs have an upward-sloping first line versus a
downward-sloping one:

```{r, eval = FALSE}
n_up <- 0; n_down <- 0
for (seed in 1:50) {
  fit <- train_network(xor_data[, c("x1","x2")], xor_data$category, n_hidden = 2, seed = seed)
  if (fit$loss_history[2000] < 0.1) {          # only count runs that actually solved XOR
    slope1 <- -fit$W1[1, 1] / fit$W1[1, 2]
    if (slope1 > 0) n_up <- n_up + 1 else n_down <- n_down + 1
  }
}
c(up = n_up, down = n_down)
```

### Exercise 4c: what does the hidden layer do to the inputs?

The output unit is just one more perceptron, so on its own it can only
draw *one* straight line, and one line cannot solve XOR. So how does the
network solve it? The trick is that the output unit doesn't look at
`(x1, x2)`. It looks at the hidden units' outputs `(h1, h2)`. Let's
compute `(h1, h2)` for each of the four XOR inputs, using the network
from Exercise 4 (seed 9):

```{r}
H <- matrix(0, nrow = 4, ncol = 2)       # one row per input pattern, one column per hidden unit
for (i in 1:4) {
  x_i <- c(xor_data$x1[i], xor_data$x2[i])
  H[i, ] <- forward(x_i, chosen_xor$W1, chosen_xor$b1, chosen_xor$w2, chosen_xor$b2)$h
}
data.frame(xor_data, h1 = round(H[, 1], 2), h2 = round(H[, 2], 2))
```

Now plot the four points in **hidden space** (`h1` on the x-axis, `h2` on
the y-axis) instead of in input space, together with the output unit's
own decision line. The output unit says "category 1" when
$v_1 h_1 + v_2 h_2 + c > 0$, so the line is where this sum equals 0, i.e.
$h_2 = -(v_1 h_1 + c)/v_2$:

```{r}
plot(H[, 1], H[, 2], pch = 19, cex = 3,
     col = ifelse(xor_data$category == 1, "steelblue", "darkorange"),
     xlab = "h1", ylab = "h2", xlim = c(-0.2, 1.2), ylim = c(-0.2, 1.2))
h1_seq <- seq(-0.2, 1.2, length.out = 100)
lines(h1_seq, -(chosen_xor$w2[1] * h1_seq + chosen_xor$b2) / chosen_xor$w2[2], lwd = 2)
```

Questions:

1. In the original input space, could *one* straight line separate the
   blue points from the orange ones? And in hidden space?
2. Look at the table of `h1` and `h2`. Which input patterns have ended
   up at (almost) the same spot in hidden space? Are they from the same
   category or different categories?
3. In your own words: what did the hidden layer do to the problem, so
   that the output unit could solve it with a single line?
4. Use this to explain Exercise 3: why can't a network with only *one*
   hidden unit solve XOR?
5. **Optional:** compare the two hidden units with the lecture's OR and
   NAND units. Which logic gate does each one resemble? (Check each
   column of the table: for which inputs is it close to 1?)

## Part 3: Does 2 hidden units scale up to 3-bit parity?

### Exercise 5

Run the exact same 10-restart pattern on `parity_data` instead of
`xor_data`:

```{r}
parity_runs <- list()
parity_final_losses <- numeric(10)
for (seed in 1:10) {
  parity_runs[[seed]] <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category, n_hidden = 2, seed = seed)
  parity_final_losses[seed] <- parity_runs[[seed]]$loss_history[2000]
}
best_parity <- parity_runs[[which.min(parity_final_losses)]]
round(parity_final_losses, 4)
best_parity$loss_history[2000]
```

How does the best final RMSD compare to Exercise 4's XOR result? Is it
clearly worse, about the same, or better? And look at the whole vector
of 10 final losses, not just the best one: how many of the 10 runs ended
up with a low RMSD?

### Exercise 6

Keeping only the best of several restarts hides how often an
*individual* run gets stuck. Here is a more direct question: if you
train a network **once**, from one random starting point, how often does
it solve parity (final RMSD below 0.15)? The function below trains
`n_runs` separate networks (seeds 1 to `n_runs`) and reports the
fraction that succeeded:

```{r}
single_run_success <- function(n_hidden, n_runs = 20) {
  successes <- logical(n_runs)           # one TRUE/FALSE per run
  for (s in 1:n_runs) {
    fit <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category,
                         n_hidden = n_hidden, seed = s)
    successes[s] <- fit$loss_history[2000] < 0.15   # TRUE if this run got below 0.15
  }
  mean(successes)                        # fraction of the runs that succeeded
}

single_run_success(n_hidden = 2)
```

What fraction of the 20 single runs succeed with `n_hidden = 2`? (This
will take a little while to run, since 20 separate networks are being
trained.) Compare with Exercise 5: how many of those 10 final RMSDs were
low?

## Part 4: How many hidden units are actually needed?

### Exercise 7

Change `n_hidden = 2` to `n_hidden = 4` in the line below and re-run
(copy the `single_run_success()` function from Exercise 6 above if it's
not still in your environment):

```{r}
single_run_success(n_hidden = 2, n_runs = 20)
```

How does the success rate change compared to Exercise 6? Try
`n_hidden = 8` too, if you're curious how far the pattern continues.

### Exercise 8 (discussion)

Based on Exercises 3, 6, and 7: does 3-bit parity need *more* hidden
units than 2-bit XOR to train reliably? Take a guess at why, in your own
words. Think back to the lecture's "two lines, combined" picture for
XOR. If XOR needs 2 lines (2 hidden units) to carve up its input space
correctly, how many lines might you guess parity-3 needs, and does that
match what you found?

## Part 5 (optional): What did the hidden units learn?

### Exercise 9 (optional)

For a successful `n_hidden = 4` run on `parity_data` (you'll need to run
`train_network()` directly, since `single_run_success()` only returned a
fraction, not the weights themselves), look at the trained `W1` matrix.
First check the run really did solve parity (final RMSD below 0.15),
then look at the weights:

```{r}
fit_4hidden <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category, n_hidden = 4, seed = 1)
fit_4hidden$loss_history[2000]   # should be below 0.15; if not, try another seed
round(fit_4hidden$W1, 2)
round(fit_4hidden$b1, 2)
```

Each row of `W1` plus its matching entry in `b1` defines one hidden
unit's own little linear rule over `(x1, x2, x3)`. Pick one hidden unit
(one row) and, using the same idea as the lecture's hand-picked OR/NAND
example, see if you can describe in words what pattern of inputs makes
that hidden unit's output close to 1 versus close to 0. Try plugging a
few different `(x1,x2,x3)` combinations into that row's weights by hand
if it helps.
