---
title: "Practical 5: Neural Networks 2 -- Solutions"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

> Posted a few days after the session. Try the exercises yourself first
> -- the value is in getting stuck, not in reading the answer.

## Setup (from the handout)

```{r}
sigmoid <- function(x) {
  1 / (1 + exp(-x))
}
forward <- function(x, W1, b1, w2, b2) {
  h <- sigmoid(W1 %*% x + b1)
  output <- sigmoid(sum(w2 * h) + b2)
  list(h = as.vector(h), output = output)
}
train_network <- function(X, y, n_hidden, seed, lr = 0.5, n_epochs = 2000) {
  set.seed(seed)
  X <- as.matrix(X)
  n_inputs <- ncol(X)
  W1 <- matrix(rnorm(n_hidden * n_inputs, sd = 1), nrow = n_hidden)
  b1 <- rep(0, n_hidden)
  w2 <- rnorm(n_hidden, sd = 1)
  b2 <- 0
  loss_history <- numeric(n_epochs)
  for (epoch in 1:n_epochs) {
    for (i in 1:nrow(X)) {
      x <- X[i, ]; target <- y[i]
      fwd <- forward(x, W1, b1, w2, b2)
      h <- fwd$h; output <- fwd$output
      delta_out <- (target - output) * output * (1 - output)
      delta_hidden <- delta_out * w2 * h * (1 - h)
      W1 <- W1 + lr * outer(delta_hidden, x)
      b1 <- b1 + lr * delta_hidden
      w2 <- w2 + lr * delta_out * h
      b2 <- b2 + lr * delta_out
    }
    predictions <- numeric(nrow(X))                      # one prediction per row of the data
    for (i in 1:nrow(X)) {
      predictions[i] <- forward(X[i, ], W1, b1, w2, b2)$output
    }
    loss_history[epoch] <- sqrt(mean((y - predictions)^2))
  }
  list(W1 = W1, b1 = b1, w2 = w2, b2 = b2, loss_history = loss_history)
}

parity_data <- expand.grid(x1 = c(0, 1), x2 = c(0, 1), x3 = c(0, 1))
parity_data$category <- (parity_data$x1 + parity_data$x2 + parity_data$x3) %% 2
```

## Exercise 1

Each row of `parity_data` should show `category = 1` exactly when one or
three of `x1`, `x2`, `x3` are 1 -- e.g. `(1,0,0)` has one 1, category 1;
`(1,1,0)` has two 1s, category 0; `(1,1,1)` has three 1s, category 1.
For the two-input case (imagining `x3` fixed at 0), this collapses to
exactly XOR.

## Exercise 2

```{r}
forward(x = c(1, 1), W1 = matrix(c(1, 1), nrow = 1), b1 = 0, w2 = 1, b2 = 0)
```

`h` comes out around 0.881, `output` around 0.707, matching the handout.
Flipping `w2` from `1` to `-1` flips `output` to about 0.293 -- the same
`h` (since `w2` doesn't affect the hidden layer), but now negatively
weighted into the output, so a confident hidden-unit signal gets turned
into a confidently *low* output instead of a confidently high one.

## Exercise 3

```{r}
x1 <- c(0, 0, 1, 1); x2 <- c(0, 1, 0, 1); category <- c(0, 1, 1, 0)
xor_data <- data.frame(x1, x2, category)

fit_1hidden <- train_network(X = xor_data[, c("x1","x2")], y = xor_data$category, n_hidden = 1, seed = 1)
fit_1hidden$loss_history[2000]
```

You should get an RMSD stuck around **0.41** -- nowhere near zero. This
matches expectations exactly: a single hidden unit, however it's
trained, is still just one perceptron computing one straight-line
boundary -- it inherits exactly the same XOR limitation from two
lectures ago. Adding a hidden *layer* only helps once there's more than
one unit in it.

## Exercise 4

```{r}
xor_runs <- list()
xor_final_losses <- numeric(10)
for (seed in 1:10) {
  xor_runs[[seed]] <- train_network(xor_data[, c("x1","x2")], xor_data$category, n_hidden = 2, seed = seed)
  xor_final_losses[seed] <- xor_runs[[seed]]$loss_history[2000]
}
round(xor_final_losses, 4)

chosen_xor <- xor_runs[[9]]          # the run from seed 9
chosen_xor$loss_history[2000]
```

The ten seeds give quite different results (some finish at a high RMSD,
clearly stuck; others reach well under 0.1) -- exactly the "some random
starts get stuck" behavior the lecture warned about, which is why
best-of-several restarts matter. The chosen run (seed 9) should land well under
0.1 (about 0.05), confirming `train_network()` reproduces the lecture's
result. We use seed 9 rather than the overall lowest RMSD because its
two lines slope downward, like the lecture's OR and NAND lines. The
boundary-line plot should show two distinct lines that, together, box
the two blue points away from the two orange ones -- the same "two
lines" picture as the lecture, just discovered automatically.

## Exercise 4b

The runs that solve XOR do **not** all find the same two lines. They
fall into two families:

- **Both lines slope downward (negative slope)**, roughly parallel to
  the anti-diagonal -- like the lecture's hand-picked OR and NAND lines
  ($x_1 + x_2 = 0.5$ and $x_1 + x_2 = 1.5$). The two blue points sit in
  the *band between* the lines, and the two orange points sit outside
  it, one on each side.
- **Both lines slope upward (positive slope)**, roughly parallel to the
  main diagonal -- lines like $x_2 = x_1 + 0.5$ and $x_2 = x_1 - 0.5$.
  Now each blue point is cut off on its own side, and the two orange
  points sit together in the band between the lines. The output unit
  fires when *either* hidden unit does.

Within a single successful run, the two lines almost always slope the
*same* way (both up or both down), and across many runs both families
turn up about equally often -- with 10 seeds you should almost always
see both, though the exact split varies. Which family a run lands in
depends entirely on its random starting weights. Runs that got stuck
(RMSD well above 0.1) show lines that don't separate the categories at
all, so ignore those. Swapping which hidden unit is "unit 1" and which
is "unit 2" also gives an equivalent solution, so even two runs in the
same family can have their lines in a different order.

So the answer to "is the solution unique?" is **no**: XOR has more than
one valid two-line solution, and gradient descent simply finds whichever
one its starting point leads to. This is true of neural networks in
general -- different training runs can end up with different weights
that behave the same way, which is one reason trained weights are hard
to interpret one by one.

## Exercise 4 (learning curves)

```{r}
plot(xor_runs[[1]]$loss_history, type = "l", lwd = 2, col = 1, ylim = c(0, 0.6),
     xlab = "epoch", ylab = "RMSD")
for (seed in 2:10) {
  lines(xor_runs[[seed]]$loss_history, lwd = 2, col = seed)
}
```

All runs start at roughly the same error (an RMSD around 0.5, i.e. the
network outputs about 0.5 for everything). After that they differ a lot.
Runs that succeed typically sit on a **plateau** near that starting error
for a while, then drop fairly suddenly to a low RMSD; how long the
plateau lasts depends on the random start. Runs that are "stuck" stay
at a high RMSD for all 2000 epochs: they have found a poor solution
(or are stuck on a long plateau) and have not made it out in the time
we gave them. This is what "stuck" means, and why restarting from a new
random start helps. (The exact curves depend on your seeds.)

## Exercise 4c

```{r}
H <- matrix(0, nrow = 4, ncol = 2)
for (i in 1:4) {
  x_i <- c(xor_data$x1[i], xor_data$x2[i])
  H[i, ] <- forward(x_i, chosen_xor$W1, chosen_xor$b1, chosen_xor$w2, chosen_xor$b2)$h
}
data.frame(xor_data, h1 = round(H[, 1], 2), h2 = round(H[, 2], 2))

plot(H[, 1], H[, 2], pch = 19, cex = 3,
     col = ifelse(xor_data$category == 1, "steelblue", "darkorange"),
     xlab = "h1", ylab = "h2", xlim = c(-0.2, 1.2), ylim = c(-0.2, 1.2))
h1_seq <- seq(-0.2, 1.2, length.out = 100)
lines(h1_seq, -(chosen_xor$w2[1] * h1_seq + chosen_xor$b2) / chosen_xor$w2[2], lwd = 2)
```

1. In input space, no single line separates the two categories (that is
   the XOR problem). In hidden space, one line does: the output unit's
   line separates the blue points from the orange ones.
2. For the seed-9 run (two downward-sloping lines, as in the lecture) the
   table should look roughly like this:

   | x1 | x2 | category | h1 | h2 |
   |----|----|----------|------|------|
   | 0 | 0 | 0 | 1.00 | 0.90 |
   | 0 | 1 | 1 | 0.87 | 0.03 |
   | 1 | 0 | 1 | 0.87 | 0.03 |
   | 1 | 1 | 0 | 0.11 | 0.00 |

   The two **blue** patterns, `(0,1)` and `(1,0)`, land at the *same* spot,
   about `(0.87, 0.03)`, while the two orange patterns land at two
   different spots on either side. `h1` is NAND-like (high except for
   `(1,1)`), as in the lecture. `h2` is high only for `(0,0)`: it is the
   *mirror image* of the lecture's OR unit (NOT OR). It draws the same
   kind of line but fires on the other side of it, and the output unit
   compensates with a negative weight on `h2`. (A hidden unit can fire
   on either side of its line; the output weight just gets the opposite
   sign to cancel it.) So the exact corner differs from the lecture's
   `(1, 1)`, but the picture is the same: blue together, orange apart.
3. The hidden layer *re-represents* the inputs: it moves the four points
   so that the two blue points are pulled together into one spot, and
   the categories can then be told apart with a single straight line.
   The output unit only has to solve an easy, linearly separable
   problem; the hard part, making XOR linearly separable, was done by the
   hidden layer.
4. With only one hidden unit, hidden space is one-dimensional: every
   input is just a single number `h1`, and a threshold on `h1` is again
   one straight line in input space. That is exactly the limitation of a
   single perceptron, so XOR cannot be solved (RMSD stays around 0.41).
5. `h1` is high for every input except `(1,1)`: that is NAND. `h2` is
   high only for `(0,0)`: that is NOR, the mirror image of the lecture's
   OR unit (see the note in question 2).

## Exercise 5

```{r}
parity_runs <- list()
parity_final_losses <- numeric(10)
for (seed in 1:10) {
  parity_runs[[seed]] <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category, n_hidden = 2, seed = seed)
  parity_final_losses[seed] <- parity_runs[[seed]]$loss_history[2000]
}
best_parity <- parity_runs[[which.min(parity_final_losses)]]
round(parity_final_losses, 4)
best_parity$loss_history[2000]
```

The best parity run should end noticeably higher than XOR's best run
(which is well under 0.1), and, more importantly, the *other* parity
runs are mostly stuck at a much higher RMSD. With only 2 hidden units,
most random starting points don't solve parity, even though a few lucky
ones do -- Exercise 6 measures exactly how often.

## Exercise 6

```{r}
single_run_success <- function(n_hidden, n_runs = 20) {
  successes <- logical(n_runs)
  for (s in 1:n_runs) {
    fit <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category,
                         n_hidden = n_hidden, seed = s)
    successes[s] <- fit$loss_history[2000] < 0.15
  }
  mean(successes)
}

single_run_success(n_hidden = 2)
```

With 2 hidden units only a **minority** of single runs succeed -- well
under half of the 20 (the exact fraction depends on seeds 1 to 20). Most
individual runs get stuck at an RMSD well above 0.15, which matches what
you saw in Exercise 5, where only a few of the 10 final losses were low.
So with 2 hidden units, parity is *possible* to learn, but a single run
usually fails -- you only succeed if you restart enough times.

## Exercise 7

```{r}
single_run_success(n_hidden = 4)
single_run_success(n_hidden = 8)
```

With `n_hidden = 4` a large majority of single runs succeed, and with
`n_hidden = 8` nearly all of them do. That's a much bigger jump than the
restart-based version showed, because here we're measuring how reliable
one run is rather than whether *some* restart eventually works. Going
from 4 to 8 helps less than going from 2 to 4 -- more hidden units help,
but with diminishing returns.

## Exercise 8 (discussion)

Yes -- 3-bit parity needs more hidden units than 2-bit XOR to train
reliably. 2 hidden units solved XOR often enough that best-of-10 always
worked (Exercise 4), but on parity a single run usually failed with 2
hidden units (Exercise 6), while with 4 hidden units most single runs
succeeded (Exercise 7). Following the "two lines, combined" picture:
XOR's four points need 2 lines to separate correctly. Parity-3 has 8
points instead of 4, arranged in a more complex pattern (a cube's
corners, not a square's) -- so it's reasonable to guess it needs *more*
than 2 lines to carve correctly, and the results back that up: 2 hidden
units are right at the edge of being enough (they only get there with
luck and restarts), while 4 is comfortably enough.

## Exercise 9 (optional)

```{r}
fit_4hidden <- train_network(parity_data[, c("x1","x2","x3")], parity_data$category, n_hidden = 4, seed = 1)
fit_4hidden$loss_history[2000]   # should be below 0.15; if not, try another seed
round(fit_4hidden$W1, 2)
round(fit_4hidden$b1, 2)
```

With seed 1 and 4 hidden units, the run solved parity (final RMSD about
0.03), and the trained weights were:

| unit | `x1` | `x2` | `x3` | bias |
|------|-------|-------|-------|-------|
| 1 | -3.90 | -1.79 | 3.12 | 1.08 |
| 2 | -2.90 | 2.50 | -0.74 | 0.29 |
| 3 | -6.69 | 6.41 | 6.63 | -3.19 |
| 4 | 2.14 | 3.97 | 3.33 | -4.74 |

The cleanest unit to read is **unit 4**. All three of its weights are
positive and its bias is clearly negative (-4.74), so it only gets above
0 when **at least two inputs are on**: with just one input on, the sum
is at most 3.97 - 4.74 < 0 (output below 0.5), with two on it is
between +0.7 and +2.6 (output about 0.67 to 0.93), and with all three on
it is +4.7 (output 0.99). So unit 4 is roughly a "two or more inputs on"
detector.

**Unit 3** is also fairly readable: it is strongly positive for `x2` and
`x3` (about +6.5 each), strongly negative for `x1` (-6.69), with bias
-3.19. It fires (output above 0.9) for `(0,0,1)`, `(0,1,0)`, `(0,1,1)`
and `(1,1,1)`, i.e. whenever `x2` or `x3` is on and `x1` is off (plus
the all-ones pattern, where `x2` and `x3` together outweigh `x1`). It
stays low (around 0.04 or lower) for `(0,0,0)`, `(1,0,0)`, `(1,0,1)` and
`(1,1,0)`.

Units 1 and 2 are messier: for example unit 1 has a negative weight on
`x1` and `x2` and a positive one on `x3`, and gives intermediate outputs
for several patterns. This is worth being honest about: unlike the
lecture's clean, hand-picked OR/NAND example, a real hidden unit trained
from a random start often learns a messier, harder-to-summarize rule, and
the network as a whole solves parity by *combining* these partial
features (some of them tidy, like "at least two inputs on", some not).
That's a normal, expected outcome of training, not a sign anything went
wrong. Your own numbers will differ with a different seed.
