---
title: "Week 3: Supervised Learning -- Solutions"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
editor_options: 
  markdown: 
    wrap: 72
---

> Posted a few days after the session. Try the exercises yourself first
> -- the value is in getting stuck, not in reading the answer.

## Setup (from the handout)

```{r}
contrast <- seq(0, 1, 0.12)
n_trials <- rep(40, 9)
n_correct <- c(16, 25, 23, 24, 25, 36, 37, 38, 40)
prop_correct <- n_correct / n_trials
data <- data.frame(contrast, n_correct, n_trials, prop_correct)

psychometric_model <- function(alpha, beta, x) {
  y <- rep(0, length(x))
  valid <- x >= 0
  y[valid] <- 0.5 + 0.5*(1 - exp(-(x[valid] / alpha)^beta))
  y
}

rmsd <- function(params, data) {
  alpha <- params[1]; beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$contrast)
  sqrt(mean((data$prop_correct - predicted)^2))
}

fit_rmsd <- optim(par = c(alpha = 0.3, beta = 1), fn = rmsd, data = data)
```

## Exercise 1

There's no single correct numeric answer -- it's an eyeballing exercise.
A reasonable guess: `alpha` somewhere around 0.5-0.6 (wherever the curve
looks like it's crossing about 75%), `beta` somewhere around 2-4 (a
moderately steep, not extremely sharp or shallow, rise).

## Exercise 2

```{r}
rmsd(c(alpha = 0.5, beta = 3), data)    # a plausible guess -- RMSD around 0.08
rmsd(c(alpha = 0.1, beta = 1), data)    # a deliberately bad guess -- RMSD around 0.24
```

The bad guess gives a noticeably higher RMSD, as expected -- roughly
three times worse. Your own Exercise 1 guess should land somewhere
closer to the first number than the second, if your eyeballing was
reasonable.

## Exercise 3

```{r}
alpha_vals <- seq(0.7, 1, length.out = 50)   # deliberately excludes the true region
beta_vals <- seq(0.1, 6, length.out = 100)
grid <- expand.grid(alpha = alpha_vals, beta = beta_vals)
grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))
best <- grid[which.min(grid$rmsd), ]
best
```

You should find `best` now reports `alpha` right at the *edge* of the
searched range (around 0.7), with an RMSD around 0.085 -- clearly worse
than the roughly 0.066 you'd get searching the full range. This is
exactly the limitation the exercise is pointing at: grid search has no
way to tell you "there's a better answer outside the box I searched" --
it just confidently reports the best point it happened to check, and a
narrowed or misplaced range gives you a wrong (if plausible-looking)
answer with no warning attached. Remember to set `alpha_vals` back to
`seq(0.01, 1, length.out = 100)` before continuing.

## Exercise 4

```{r}
fit_far <- optim(par = c(alpha = 0.9, beta = 0.2), fn = rmsd, data = data)
fit_far$par
```

You should get essentially the same answer as `fit_rmsd$par` (around
`alpha ≈ 0.585, beta ≈ 3.25`), despite starting quite far from it in
both parameters. Try a couple more starting points of your own -- for
this particular dataset, you should find `optim()` reliably lands in the
same place almost regardless of where it starts, because there's only
one clear valley in the RMSD surface (as you'll see explicitly if you
ever plot the full cost surface).

## Exercise 5

There's no single expected answer here since it depends on which values
you tried, but the general pattern: recovery works well for most
reasonable combinations. It breaks down specifically with a very small
`true_alpha`, such as 0.05:

```{r}
set.seed(99)
contrast_sim <- seq(0, 1, 0.12)
true_alpha <- 0.05
true_beta  <- 2.5
p_true <- psychometric_model(true_alpha, true_beta, contrast_sim)
n_trials_sim <- rep(40, 9)
n_correct_sim <- rbinom(length(contrast_sim), n_trials_sim, p_true)
prop_correct_sim <- n_correct_sim / n_trials_sim

data_sim <- data.frame(contrast = contrast_sim, n_correct = n_correct_sim,
                        n_trials = n_trials_sim, prop_correct = prop_correct_sim)
data_sim

fit_sim <- optim(par = c(alpha = 0.3, beta = 1), fn = rmsd, data = data_sim)
fit_sim$par
```

With `true_alpha = 0.05`, every contrast level except `contrast = 0` is
already at (or essentially at) ceiling performance in the simulated data
-- there's really only *one* informative data point left (the one at
`contrast = 0`), since all the others just say "yes, definitely correct"
regardless of the true curve's exact shape past that. With that little
information, `optim()` typically finds a fit that's an *excellent* match
to this particular simulated dataset -- often an even lower RMSD than
the true parameters would give -- while landing on `alpha`/`beta` values
nowhere near the true `(0.05, 2.5)`. That's a genuinely important
result: a great RMSD doesn't guarantee you've recovered the true
parameters, especially when the data itself doesn't contain enough
information to distinguish between many different curves. Remember to
set `true_alpha`/`true_beta` back to `0.5`/`2.5` before continuing.

## Exercise 6 (interpretation)

Comparing `alpha ≈ 0.49` (the `true_alpha = 0.5` simulated participant)
to `alpha ≈ 0.59` (the real Part 0-4 participant, `fit_rmsd$par`): the
simulated participant has a **lower threshold** -- they reach a given
detection level at lower contrast, so they're more sensitive overall.
The simulated participant's `beta` (around 2.5) is a bit lower than the
real participant's (around 3.25), meaning their rise is slightly more
gradual -- somewhat less consistent responding right around their own
threshold. These are two separate psychological dimensions (overall
sensitivity vs. reliability of responding), and a participant can score
differently on each.

## Exercise 7

```{r}
linear_model <- function(a, b, x) {
  a + b * x
}
rmsd_linear <- function(params, data) {
  a <- params[1]; b <- params[2]
  predicted <- linear_model(a, b, data$contrast)
  sqrt(mean((data$prop_correct - predicted)^2))
}
fit_linear <- optim(par = c(a = 0.5, b = 0.5), fn = rmsd_linear, data = data)

linear_model(fit_linear$par[1], fit_linear$par[2], 0)      # about 0.44
linear_model(fit_linear$par[1], fit_linear$par[2], 1)      # about 1.05
linear_model(fit_linear$par[1], fit_linear$par[2], 1.5)    # about 1.35
linear_model(fit_linear$par[1], fit_linear$par[2], -0.5)   # about 0.14
```

At `contrast = 1`, the linear model already predicts *above* 1.0 -- more
than 100% correct, which is impossible. Pushed further out to
`contrast = 1.5`, it gets even more extreme (about 1.35). There's
nothing stopping a straight line from continuing past 1 or below 0; it
has no concept of a ceiling or floor.

## Exercise 8 (discussion)

```{r}
psychometric_model(fit_rmsd$par[1], fit_rmsd$par[2], c(0, 1))
```

Yes -- the linear model's Exercise 7 prediction at `contrast = 1` (about
1.05) is outside the range a proportion can actually take, and the same
problem shows up (even more clearly) once you push past the tested range
entirely, as at `contrast = 1.5`. The Weibull model, evaluated at the
same two points, gives `0.5` at `contrast = 0` and something just under
`1.0` at `contrast = 1` -- both perfectly valid proportions. The Weibull
model can never do what the linear model just did, because it's built
from a sigmoid: mathematically, `0.5 + 0.5*(1 - exp(-(x/alpha)^beta))`
can only ever approach 0.5 or 1 as limits, no matter how extreme `x`
gets -- it's structurally incapable of predicting an impossible
proportion, which is exactly why the shape of a model, not just how well
it fits the tested range, matters.

## Exercise 9

```{r}
time_days <- c(0, 1, 2, 4, 7, 14, 21, 30)
prop_recalled <- c(0.96, 0.88, 0.85, 0.76, 0.57, 0.35, 0.27, 0.14)
forgetting_data <- data.frame(time_days, prop_recalled)

forgetting_model <- function(a, b, t) {
  a * exp(-b * t)
}
rmsd_forgetting <- function(params, data) {
  a <- params[1]; b <- params[2]
  predicted <- forgetting_model(a, b, data$time_days)
  sqrt(mean((data$prop_recalled - predicted)^2))
}
fit_forgetting <- optim(par = c(a = 0.9, b = 0.05), fn = rmsd_forgetting, data = forgetting_data)

t_half <- log(fit_forgetting$par[1] / 0.5) / fit_forgetting$par[2]
t_half
```

You should get something close to **9.8 days** -- the "half-life" of
this simulated memory. Checking against the data plot: recall is 0.57 at
day 7 and 0.35 at day 14, so 0.5 genuinely falls between those two
points -- 9.8 days is right where you'd expect it to land.

## Exercise 10 (discussion)

With the original target of 0.8:

```{r}
t_08 <- log(fit_forgetting$par[1] / 0.8) / fit_forgetting$par[2]
t_08
```

You should get close to **2.7 days**. Checking against the data: recall
is 0.85 at day 2 and 0.76 at day 4, so 0.8 genuinely falls between those
two points -- a good sanity check that the fitted curve and the
inference built on it are both behaving sensibly.

If the fitted `a` had come out *below* the target proportion you're
solving for: the formula would still compute a number (there's no
division by zero or anything that literally breaks), but that number
would come out *negative*. That's the model's way of saying the target
proportion was already crossed *before* time zero -- a crossing point in
the past, before learning even happened, which isn't a meaningful answer
for a real forgetting curve. This is a useful general lesson: an
inference computed from a fitted model is only as meaningful as whether
the question you asked actually makes sense given the fit you got -- the
algebra will still hand you a number even once the underlying question
has stopped making sense, which is exactly why sanity-checking against
the real data, like you just did, matters rather than trusting a formula
blindly.
