---
title: "Practical 1: R Exercises I -- Solutions"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
editor_options: 
  markdown: 
    wrap: 72
---

> Try the exercises yourself first -- the value is in getting stuck, not
> in reading the answer.

------------------------------------------------------------------------

## Part 1: Variables and basic operations

### Exercise 1

```{r}
rt1 <- 452
rt2 <- 618
rt3 <- 390
average_rt <- (rt1 + rt2 + rt3) / 3
average_rt
```

### Exercise 2

```{r}
score <- 0
score <- score + 4
score
score <- score - 1
score
score <- score + 2
score
```

### Exercise 3

```{r}
r <- 7.5
area <- pi * r^2
area
```

------------------------------------------------------------------------

## Part 2: Data types

### Exercise 4

```{r}
age_text <- "25"
# age_text * 2  -- this line would error: non-numeric argument to binary operator.
# "25" is a character string, not a number -- R won't silently guess you meant the number 25.

age_number <- as.numeric(age_text)
age_number * 2
```

### Exercise 5

```{r}
5 > 3        # TRUE
5 == "5"     # TRUE -- R converts 5 to the character "5" to compare them, since one side is already text
!(5 > 3)     # FALSE
"cat" == "dog"  # FALSE
```

The `5 == "5"` result is the one worth pausing on: it's `TRUE`, not an
error and not `FALSE`. R quietly converts the number to text before
comparing, rather than complaining about the type mismatch -- exactly
the kind of "quiet conversion" the manual warns about.

### Exercise 6

```{r}
responded_correctly <- TRUE
ifelse(responded_correctly, "correct", "incorrect")
```

------------------------------------------------------------------------

## Part 3: Numerical gotchas

### Exercise 7

```{r}
1 / 0      # Inf
-1 / 0     # -Inf
0 / 0      # NaN
log(0)     # -Inf
log(-1)    # NaN, with a warning
```

### Exercise 8

```{r}
test_values <- list(NA, NaN, Inf, 5)
for (v in test_values) {
  cat(sprintf("value: %-6s  is.na: %-6s  is.nan: %-6s  is.finite: %-6s\n",
              as.character(v), is.na(v), is.nan(v), is.finite(v)))
}
```

You should find `is.na(NaN)` comes out `TRUE` -- genuinely surprising
the first time you see it. Looking up `?is.na` explains why: R's help
page states explicitly that `is.na()` is also `TRUE` for `NaN`, because
`NaN` is implemented internally as a special case of the missing-value
family, not as something wholly separate. `is.nan()` is the stricter
check if you specifically want to know "is this exactly NaN, not just
any kind of missing/undefined value."

### Exercise 9

```{r}
values <- c(3, 5, NaN, 2, 8)
sum(values)
mean(values)
```

Both come out `NaN`, even though only one of the five values is actually
the problem. `sum()` and `mean()` don't skip over the bad value and work
around it by default -- one `NaN` anywhere in the input makes the whole
result `NaN`. (If you're curious: both functions do have an
`na.rm = TRUE` argument that would skip missing/NaN values instead --
worth knowing exists, even though the point of this exercise is to see
the default "poisoning" behavior first.)

------------------------------------------------------------------------

## Part 4: Calling functions

### Exercise 10

```{r}
round(3.14159, 2)
```

### Exercise 11

```{r}
rnorm(8, mean = 100, sd = 15)
```

(Your actual numbers will differ each time you run this -- `rnorm()`
draws new random values every call, unless you fix them with
`set.seed()` first, which you'll meet properly in a later practical.)

### Exercise 12

```{r}
sqrt(log(50))
```

------------------------------------------------------------------------

## Part 5: Vectors

### Exercise 13

```{r}
reaction_times <- c(452, 618, 390, 705, 511)
```

### Exercise 14

```{r}
mean(reaction_times)
max(reaction_times)
reaction_times - mean(reaction_times)
```

### Exercise 15

```{r}
reaction_times[c(2, 4)]                          # (a)
reaction_times[reaction_times > 500]              # (b)
reaction_times[-which.max(reaction_times)]        # (c)
```

### Exercise 16

```{r}
ages <- c(Alex = 28, Priya = 34, Sam = 25)
ages["Priya"]
```

### Exercise 17

```{r}
seq(0, 1, length.out = 10)
```

### Exercise 18 (discussion)

R applies `-` between a vector and a single number by **recycling** the
single number against every element of the vector -- effectively
treating `mean(reaction_times)` as if it were repeated once for every
element, then subtracting position by position. That's the same
"operations apply to every element at once" rule from the manual's
coverage of vectors, just for subtraction specifically rather than
addition or multiplication. Behind the scenes this is doing exactly the
same work a for-loop would do, one element at a time -- next session's
final section makes that equivalence completely explicit, with a real
for-loop and its vectorized equivalent side by side.
