---
title: "Week 3: Supervised Learning \ Parameter Estimation"
subtitle:  "Basic Methods & Maximum Likelihood"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
---

## Before you start

In the lecture you fit a Weibull-based psychometric function to 2AFC contrast-detection data using both RMSD and maximum likelihood. This practical has you rebuild the whole pipeline yourself, then use it in three different ways: recovering known parameters from simulated data, comparing the Weibull model against a plain linear model, and — a new, optional section at the end — fitting a completely different psychometric curve (a forgetting curve) and using the fitted model to answer a real question, not just report a fit.

You don't need to understand every line of R syntax perfectly — focus on understanding *what each step is doing conceptually*. If something errors, re-check you've run all the chunks above it in order.

## Part 0: The data

Same 2AFC contrast-detection setup as the lecture: on each trial, a participant says which of two intervals contained the stimulus, at one of nine contrast levels, 40 trials per level.

```{r}
contrast <- seq(0, 1, 0.12)
n_trials <- rep(40, 9)
n_correct <- c(16, 25, 23, 24, 25, 36, 37, 38, 40)
prop_correct <- n_correct / n_trials

data <- data.frame(contrast, n_correct, n_trials, prop_correct)
data
```

```{r}
old_par <- par(pty = "s")
plot(data$contrast, data$prop_correct, pch = 19,
     xlim = c(0, 1), ylim = c(0.4, 1),
     xlab = "Stimulus contrast", ylab = "Proportion correct",
     main = "Psychometric function")
par(old_par)
```

## Part 1: A first guess, by eye

### Exercise 1

Looking at the plot above, write down your own rough guess for `alpha` and `beta` — where does the curve look like it crosses roughly 75% correct (`alpha`), and how steep does the rise look (`beta`)? There's no single right answer here; this is just a benchmark to compare against later.

## Part 2: How wrong is a guess? — RMSD

### Exercise 2 

Copy in `psychometric_model()` and `rmsd()` from the lecture (unchanged), then try your Exercise 1 guess alongside a couple of others.

```{r}
# Your code here:
# - psychometric_model <- function(alpha, beta, x) { ... } 
# - rmsd <- function(params, data) { ... }
# - try rmsd() on your own guess and 1-2 others, using data$contrast and data$prop_correct

```

## Part 3: Grid search

### Exercise 3 

Build a grid of candidate `(alpha, beta)` pairs and find the best-fitting combination. `alpha` should range across roughly the same interval as `contrast` itself (contrast only goes up to 1, so there's no reason to search alpha values far outside that range); `beta` can range more widely.

*Hint:* `alpha_vals <- seq(0.01, 1, length.out = 100)` (starting just above 0, not exactly 0, to avoid a division-by-zero inside the model) and `beta_vals <- seq(0.1, 6, length.out = 100)` are reasonable ranges to start from.

```{r}
# Your code here:
# - alpha_vals, beta_vals, grid <- expand.grid(...)
# - grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))
# - best <- grid[which.min(grid$rmsd), ]

```

## Part 4: Automatic search with `optim()`

### Exercise 4

Run `optim()` from a deliberately poor starting point — something like `alpha = 0.02, beta = 5.5` — and confirm it still finds essentially the same answer as your grid search.

```{r}
# Your code here

```

### Exercise 5

Now run it from a more reasonable starting point (e.g. `alpha = 0.3, beta = 1`) and save the result as `fit_rmsd` — you'll need this later in the practical.

```{r}
# Your code here

```

## Part 5: Maximum likelihood

### Exercise 6

Write `nll()` (negative log-likelihood), matching the lecture's version — remember to clamp the predicted probability away from exactly 0 or 1 before passing it to `dbinom()`. Fit it with `optim()`, starting from the same point as Exercise 5, and save the result as `fit_mle`.

```{r}
# Your code here:
# - nll <- function(params, data) { ... } with predicted <- pmin(pmax(predicted, 1e-6), 1 - 1e-6)
# - fit_mle <- optim(par = c(alpha = 0.3, beta = 1), fn = nll, data = data)

```

### Exercise 7 (Discussion) 

Compare `fit_rmsd$par` and `fit_mle$par`. Then think about *why* they might differ, even slightly: RMSD treats a miss of a given size the same everywhere on the curve, while likelihood treats a miss near the edges of the range (close to 0.5 or close to 1) very differently from the same-sized miss near the middle. Can you think of a situation where this difference would actually matter for which model you'd trust more?

## Part 6: Simulating your own data

A dataset you didn't generate yourself can only tell you whether your fitting method finds *some* consistent minimum — not whether that minimum is actually the *true* parameters. Simulating your own data, with known "true" values, lets you check that.

```{r,  eval=FALSE}
set.seed(99)
contrast_sim <- seq(0, 1, 0.12)
true_alpha <- 0.5
true_beta  <- 2.5
p_true <- psychometric_model(true_alpha, true_beta, contrast_sim)

n_trials_sim <- rep(40, 9)
n_correct_sim <- rbinom(length(contrast_sim), n_trials_sim, p_true)
prop_correct_sim <- n_correct_sim / n_trials_sim

data_sim <- data.frame(contrast = contrast_sim, n_correct = n_correct_sim,
                        n_trials = n_trials_sim, prop_correct = prop_correct_sim)
data_sim
```

### Exercise 8 

Fit `data_sim` with `optim()` and RMSD, starting from `alpha = 0.3, beta = 1` as before.

```{r}
# Your code here

```

### Exercise 9 

Compare your fitted `(alpha, beta)` to `true_alpha`/`true_beta`. How close did you get?

```{r}
# Your code here

```

### Exercise 10 

Plot `data_sim`'s points, and overlay your fitted curve.

```{r}
# Your code here:
# - plot(data_sim$contrast, data_sim$prop_correct, ...)
# - x_seq <- seq(0, 1, length.out = 200)
# - lines(x_seq, psychometric_model(fit_sim$par[1], fit_sim$par[2], x_seq), col = "blue")

```

### Exercise 11 (Interpretation) 

Compare this simulated participant's fitted `alpha` to the real Part 0–5 participant's `fit_rmsd$par`. Which participant is more *sensitive* overall (reaches a given detection level at lower contrast)? Which participant's responses are more *reliable* around their own threshold (steeper `beta`)? These are two genuinely separate psychological questions — a participant can be sensitive but inconsistent, or insensitive but very reliable in their own way.

### Exercise 12 (extension)

Try re-running the simulation with a few different `true_alpha`/`true_beta` combinations (edit the `set.seed()` / true value lines above and re-run Exercises 8–9). Does recovery still work well for every combination you try? Look specifically at what happens with a very small `true_alpha` (e.g. 0.05) — what does that do to the data, and why might that make the fit less reliable?

## Part 7: Does the shape of the model matter?

### Exercise 13

Write `linear_model(a, b, x)`, returning `a + b * x`.

```{r}
# Your code here

```

### Exercise 14 

Write `rmsd_linear()`, analogous to `rmsd()` but using `linear_model()`.

```{r}
# Your code here

```

### Exercise 15 

Fit the linear model to the original `data` (not `data_sim`) with `optim()`. A reasonable starting point is `a = 0.5, b = 0.5`.

```{r}
# Your code here

```

### Exercise 16 

Compare `fit_linear$value` to `fit_rmsd$value`. Which model fits better by this measure?

```{r}
# Your code here

```

### Exercise 17 

Plot both fitted curves over the data, extending `ylim` a bit beyond $$0, 1$$ so any out-of-bounds predictions are visible (e.g. `ylim = c(-0.2, 1.2)`).

```{r}
# Your code here

```

### Exercise 18 (discussion)

Check the linear model's prediction at `contrast = 0` and `contrast = 1`. Is either prediction outside the range a proportion can actually take? What does the Weibull model do instead at those same points, and why can it never do what the linear model just did?

## Part 8: The forgetting curve — fit it, then use it

Everything so far has been about detecting a stimulus. This last part applies the exact same fitting machinery to a completely different psychological phenomenon: how memory for a learned item decays over time — the classic **forgetting curve**.

```{r}
time_days <- c(0, 1, 2, 4, 7, 14, 21, 30)
prop_recalled <- c(0.96, 0.88, 0.85, 0.76, 0.57, 0.35, 0.27, 0.14)

forgetting_data <- data.frame(time_days, prop_recalled)
forgetting_data
```

### Exercise 19 

Plot `prop_recalled` against `time_days`. What shape is this — does it look like it could be described by the same S-shaped Weibull curve from Part 0–7, or does it look different?

```{r}
# Your code here

```

### Exercise 20 

We'll model this with a classic two-parameter exponential forgetting curve:

$$p(t) = a \cdot e^{-b \cdot t}$$

Here `a` is the (estimated) proportion recalled at time zero, right after learning, and `b` is the decay rate — how quickly memory fades. Write this as `forgetting_model(a, b, t)`.

```{r}
# Your code here

```

### Exercise 21 

Write `rmsd_forgetting()`, analogous to `rmsd()` from Part 2, and fit it with `optim()`. A reasonable starting point is `a = 0.9, b = 0.05`.

```{r}
# Your code here

```

### Exercise 22 

Here's the actual inference question: **how many days pass before recall drops to 0.8 (80% correct)?** Your fitted model gives you `a` and `b` — you're not given this answer directly, you have to derive it. Setting $p(t) = 0.8$ and solving for $t$:

$$0.8 = a \cdot e^{-bt} \quad\Longrightarrow\quad t = \frac{\ln(a / 0.8)}{b}$$

Compute this using your fitted `a` and `b`.

```{r}
# Your code here:
# - t_08 <- log(fit_forgetting$par[1] / 0.8) / fit_forgetting$par[2]

```

### Exercise 23 (discussion)
Sanity-check your answer to Exercise 22 against the raw data plot from Exercise 19 — does the time you calculated fall in a sensible place relative to where the observed points cross 0.8? What would it mean, mathematically, if your fitted `a` had come out *below* 0.8? (Think about what the formula in Exercise 22 would do in that case, and whether that outcome would make sense for a real forgetting curve.)
