---
title: "Practical 3: Supervised Learning"
subtitle: "Basic Methods for Fitting a Model to Data"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
editor_options: 
  markdown: 
    wrap: 72
---

## Before you start

In the lecture you fit a Weibull-based psychometric function to 2AFC
contrast-detection data using RMSD. This practical has you use that same
pipeline in three different ways: recovering known parameters from
simulated data, comparing the Weibull model against a plain linear
model, and fitting a completely different psychometric curve (a
forgetting curve) and using the fitted model to answer a real question,
not just report a fit.

Most of the code below is already written for you. Your job in each
exercise is usually to **run it, then change something** — a parameter
value, a data value, a starting point — and see what happens. You don't
need to understand every line of R syntax perfectly; focus on
understanding *what changing that one value actually does*. If something
errors, re-check you've run all the chunks above it in order.

## Part 0: The data

Same 2AFC contrast-detection setup as the lecture: on each trial, a
participant says which of two intervals contained the stimulus, at one
of nine contrast levels, 40 trials per level.

```{r}
contrast <- seq(0, 1, 0.12)
n_trials <- rep(40, 9)
n_correct <- c(16, 25, 23, 24, 25, 36, 37, 38, 40)
prop_correct <- n_correct / n_trials

data <- data.frame(contrast, n_correct, n_trials, prop_correct)
data
```

```{r}
old_par <- par(pty = "s")
plot(data$contrast, data$prop_correct, pch = 19,
     xlim = c(0, 1), ylim = c(0.4, 1),
     xlab = "Stimulus contrast", ylab = "Proportion correct",
     main = "Psychometric function")
par(old_par)
```

## Part 1: A first guess, by eye

### Exercise 1

Looking at the plot above, write down your own rough guess for `alpha`
and `beta` — where does the curve look like it crosses roughly 75%
correct (`alpha`), and how steep does the rise look (`beta`)? There's no
single right answer here; this is just a benchmark to compare against
later.

## Part 2: How wrong is a guess? — RMSD

Here are `psychometric_model()` and `rmsd()`, exactly as covered in the
lecture:

```{r}
psychometric_model <- function(alpha, beta, x) {
  y <- rep(0, length(x))
  valid <- x >= 0
  y[valid] <- 0.5 + 0.5*(1 - exp(-(x[valid] / alpha)^beta))
  y
}

rmsd <- function(params, data) {
  alpha <- params[1]; beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$contrast)
  sqrt(mean((data$prop_correct - predicted)^2))
}
```

### Exercise 2

Try `rmsd()` on your own Exercise 1 guess:

```{r}
rmsd(c(alpha = 0.5, beta = 3), data)   # replace 0.5 and 3 with your own guess
```

Now try a few more combinations by changing the numbers above and
re-running — including at least one combination you're confident is a
*bad* fit. Does `rmsd()` give a clearly higher number for it, as you'd
expect?

## Part 3: Grid search

Here's a grid search over candidate `(alpha, beta)` pairs:

```{r}
alpha_vals <- seq(0.01, 1, length.out = 100)
beta_vals <- seq(0.1, 6, length.out = 100)
grid <- expand.grid(alpha = alpha_vals, beta = beta_vals)
grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))
best <- grid[which.min(grid$rmsd), ]
best
```

Here's what that whole grid looks like, all 10,000 combinations at once
— darker means a better (lower-RMSD) fit:

```{r}
rmsd_matrix <- matrix(grid$rmsd, nrow = length(alpha_vals), ncol = length(beta_vals))
image(alpha_vals, beta_vals, rmsd_matrix, xlab = "alpha", ylab = "beta",
      main = "RMSD across the grid", col = hcl.colors(50, "Blues", rev = TRUE))
contour(alpha_vals, beta_vals, rmsd_matrix, add = TRUE, nlevels = 8)
points(best$alpha, best$beta, pch = 4, col = "red", lwd = 2, cex = 1.5)
```

The red X marks `best` — the single darkest point, exactly where the
contour lines close in around a minimum. This is the same "cost surface"
idea from the lecture, just for this dataset's own grid.

### Exercise 3

The range searched matters. Change `alpha_vals` above to
`seq(0.7, 1, length.out = 50)` — a range that deliberately excludes
where you'd expect the true best-fitting `alpha` to be — and re-run the
whole chunk. What does `best` give you now? Is the reported RMSD better
or worse than before? What does this tell you about grid search's
biggest limitation: it can only ever report the best point *within the
range you gave it*, with no way to warn you that a better answer might
exist outside that range. (Change `alpha_vals` back to the original
`seq(0.01, 1, length.out = 100)` before moving on.)

## Part 4: Automatic search with `optim()`

```{r}
fit_rmsd <- optim(par = c(alpha = 0.02, beta = 5.5), fn = rmsd, data = data)
fit_rmsd$par
fit_rmsd$value
```

### Exercise 4

Change the starting point above — `par = c(alpha = 0.02, beta = 5.5)` —
to a couple of other values of your choosing, and re-run. Does `optim()`
still land on essentially the same answer every time? This particular
starting point is a deliberately extreme one — the model's predictions
are already flattened out to near-1.0 everywhere except `contrast = 0` —
and it's worth noting `optim()` still finds its way to the right answer
despite that. (Leave the chunk on this starting point, saved as
`fit_rmsd`, before moving on — you'll need it later in the practical.)

## Part 5: Simulating your own data

A dataset you didn't generate yourself can only tell you whether your
fitting method finds *some* consistent minimum — not whether that
minimum is actually the *true* parameters. Simulating your own data,
with known "true" values, lets you check that.

```{r}
set.seed(99)
contrast_sim <- seq(0, 1, 0.12)
true_alpha <- 0.5
true_beta  <- 2.5
p_true <- psychometric_model(true_alpha, true_beta, contrast_sim)

n_trials_sim <- rep(40, 9)
n_correct_sim <- rbinom(length(contrast_sim), n_trials_sim, p_true)
prop_correct_sim <- n_correct_sim / n_trials_sim

data_sim <- data.frame(contrast = contrast_sim, n_correct = n_correct_sim,
                        n_trials = n_trials_sim, prop_correct = prop_correct_sim)
data_sim
```

```{r}
fit_sim <- optim(par = c(alpha = 0.3, beta = 1), fn = rmsd, data = data_sim)
fit_sim$par                # compare this to true_alpha / true_beta above
c(true_alpha = true_alpha, true_beta = true_beta)

x_seq <- seq(0, 1, length.out = 200)
plot(data_sim$contrast, data_sim$prop_correct, pch = 19,
     xlab = "Contrast", ylab = "Proportion correct", ylim = c(0, 1))
lines(x_seq, psychometric_model(fit_sim$par[1], fit_sim$par[2], x_seq), col = "blue")
```

### Exercise 5

Change `true_alpha` and `true_beta` above to a few different
combinations of your own choosing, and re-run both chunks each time (the
`set.seed(99)` line keeps the random noise reproducible, so you can
compare fairly). Does `fit_sim$par` stay close to your new true values
every time? Now specifically try a very small `true_alpha`, like `0.05`
— what happens to the shape of the simulated data, and does recovery
still work as well? (Set `true_alpha` and `true_beta` back to `0.5` and
`2.5` before moving on.)

### Exercise 6 (interpretation)

With `true_alpha = 0.5, true_beta = 2.5` restored, compare this
simulated participant's fitted `alpha` to the real Part 0–4
participant's `fit_rmsd$par`. Which participant is more *sensitive*
overall (reaches a given detection level at lower contrast)? Which
participant's responses are more *reliable* around their own threshold
(steeper `beta`)? These are two genuinely separate psychological
questions — a participant can be sensitive but inconsistent, or
insensitive but very reliable in their own way.

## Part 6: Does the shape of the model matter?

Here's a plain linear model fit to the same data, alongside the Weibull
fit from Part 4:

```{r}
linear_model <- function(a, b, x) {
  a + b * x
}

rmsd_linear <- function(params, data) {
  a <- params[1]; b <- params[2]
  predicted <- linear_model(a, b, data$contrast)
  sqrt(mean((data$prop_correct - predicted)^2))
}

fit_linear <- optim(par = c(a = 0.5, b = 0.5), fn = rmsd_linear, data = data)
fit_linear$par
fit_linear$value
fit_rmsd$value
```

```{r}
plot(data$contrast, data$prop_correct, pch = 19, ylim = c(-0.2, 1.2),
     xlab = "Contrast", ylab = "Proportion correct")

x_seq <- seq(0, 1, length.out = 200)
lines(x_seq, psychometric_model(fit_rmsd$par[1], fit_rmsd$par[2], x_seq), col = "darkgreen")
lines(x_seq, linear_model(fit_linear$par[1], fit_linear$par[2], x_seq), col = "blue")
abline(h = c(0, 1), lty = "dotted")
legend("topleft", legend = c("Weibull fit", "Linear fit"),
       col = c("darkgreen", "blue"), lty = 1)
```

### Exercise 7

Using `linear_model()` and `fit_linear$par` above, compute the linear
model's predicted proportion correct at `contrast = 0` and at
`contrast = 1`:

```{r}
linear_model(fit_linear$par[1], fit_linear$par[2], 0)
linear_model(fit_linear$par[1], fit_linear$par[2], 1.5)
```

Now try `contrast = 1.5` and `contrast = -0.5` — values outside anything
actually tested. What do you notice about these predicted values
compared to what a real proportion can be?

### Exercise 8 (discussion)

Is either of the Exercise 7 predictions at `contrast = 0` or
`contrast = 1` outside the range a proportion can actually take? What
does the Weibull model do instead at those same two points (try
`psychometric_model(fit_rmsd$par[1], fit_rmsd$par[2], c(0, 1))`
yourself), and why can it never do what the linear model just did?

## Part 7: The forgetting curve — fit it, then use it

Everything so far has been about detecting a stimulus. This last part
applies the exact same fitting machinery to a completely different
psychological phenomenon: how memory for a learned item decays over time
— the classic **forgetting curve**.

```{r}
time_days <- c(0, 1, 2, 4, 7, 14, 21, 30)
prop_recalled <- c(0.96, 0.88, 0.85, 0.76, 0.57, 0.35, 0.27, 0.14)
forgetting_data <- data.frame(time_days, prop_recalled)

plot(forgetting_data$time_days, forgetting_data$prop_recalled, pch = 19,
     xlab = "Days since learning", ylab = "Proportion recalled", ylim = c(0, 1))
```

We'll model this with a classic two-parameter exponential forgetting
curve, $p(t) = a \cdot e^{-b \cdot t}$, where `a` is the (estimated)
proportion recalled at time zero and `b` is the decay rate:

```{r}
forgetting_model <- function(a, b, t) {
  a * exp(-b * t)
}

rmsd_forgetting <- function(params, data) {
  a <- params[1]; b <- params[2]
  predicted <- forgetting_model(a, b, data$time_days)
  sqrt(mean((data$prop_recalled - predicted)^2))
}

fit_forgetting <- optim(par = c(a = 0.9, b = 0.05), fn = rmsd_forgetting, data = forgetting_data)
fit_forgetting$par
```

```{r}
plot(forgetting_data$time_days, forgetting_data$prop_recalled, pch = 19,
     xlab = "Days since learning", ylab = "Proportion recalled", ylim = c(0, 1))

t_seq <- seq(0, 30, length.out = 200)
lines(t_seq, forgetting_model(fit_forgetting$par[1], fit_forgetting$par[2], t_seq), col = "blue")
```

Here's the actual inference question: **how many days pass before recall
drops to 0.8 (80%)?** Setting $p(t) = 0.8$ and solving for $t$ gives
$t = \ln(a/0.8)/b$:

```{r}
t_08 <- log(fit_forgetting$par[1] / 0.8) / fit_forgetting$par[2]
t_08
```

### Exercise 9

Change the `0.8` in the formula above to `0.5` instead — the time until
recall drops to *half* its initial level, sometimes called a
"half-life." Recompute `t_08` (or rename it if you like) with this new
target. Does the resulting number of days make sense relative to the
data plot above?

### Exercise 10 (discussion)

With the original target of 0.8: sanity-check your Exercise-9-style
answer against the raw data plot — does the time you calculated fall in
a sensible place relative to where the observed points cross that
target? What would it mean, mathematically, if your fitted `a` had come
out *below* the target proportion you were solving for? (Think about
what the formula would do in that case, and whether the result would
make sense for a real forgetting curve.)
