---
title: "Week 4: Neural Networks 1"
subtitle: "Generalizing the Perceptron and the Delta Rule"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
---

## Before you start

In the lecture you built a two-input perceptron (`w1`, `w2`, `bias`) on the roundness/elongation shape data,
trained it with the delta rule, and watched it hit a wall on XOR. Rather than rebuilding the same two-input
version here, this practical pushes it one step further: you'll generalize your perceptron to handle *any*
number of inputs using vectors, apply it to a shape dataset with a third feature, and then bring it back down
to two inputs at the very end to meet the XOR wall yourself.

This matters beyond just "more practice": real networks don't have two hardcoded inputs named `w1` and `w2`
— they have a *vector* of weights, one per input, however many there are. Writing `perceptron()` the vector
way here is a small, concrete step toward the kind of code you'll see when hidden layers show up next week.

You don't need to understand every line of R syntax perfectly — focus on understanding *what each step is
doing conceptually*. If something errors, re-check you've run all the chunks above it in order.

## Part 0: A three-feature shape dataset

Same idea as the lecture's shapes, but this time each shape is described by **three** numbers instead of two:
roundness, elongation, and symmetry.

```{r}
roundness  <- c(0.9, 0.85, 0.8, 0.75, 0.7, 0.65, 0.6, 0.3,
                0.2, 0.15, 0.25, 0.3, 0.35, 0.4, 0.45, 0.7)
elongation <- c(0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.75,
                0.7, 0.8, 0.75, 0.65, 0.6, 0.55, 0.5, 0.2)
symmetry   <- c(0.82, 0.78, 0.88, 0.74, 0.69, 0.64, 0.58, 0.32,
                0.28, 0.22, 0.30, 0.36, 0.42, 0.46, 0.50, 0.60)
category   <- c(1, 1, 1, 1, 1, 1, 1, 1,
                0, 0, 0, 0, 0, 0, 0, 0)

data <- data.frame(roundness, elongation, symmetry, category)
data
```

**Question 1.** With three features instead of two, we can no longer plot the data on a single 2D scatter
plot the way the lecture did. Make two separate scatter plots instead — roundness vs. elongation, and
roundness vs. symmetry (colour the points by `category` in both, as in the lecture) — and describe in a
sentence or two whether symmetry looks like it adds useful information on top of roundness and elongation,
or whether it looks mostly redundant with them.

```{r}
# Your two plots here

```

**Question 1b (optional).** Two 2D plots just about work for 3 features, but that trick breaks down fast —
with 10 or 50 features you can't look at every pair. One common fix is **dimensionality reduction**:
squashing many features down into 2 numbers that preserve as much of the original structure as possible, so
you can still plot it. Base R can do a simple version of this with no extra packages, using **Principal
Component Analysis**:

```{r}
pca <- prcomp(data[, c("roundness", "elongation", "symmetry")], scale. = TRUE)
plot(pca$x[, 1], pca$x[, 2], pch = 19,
     col = ifelse(data$category == 1, "steelblue", "darkorange"),
     xlab = "PC1", ylab = "PC2")
```

Do the two categories still separate cleanly in this compressed view? *(You may see the name* **t-SNE**
*elsewhere as a fancier alternative to PCA — it tends to preserve local neighborhood structure better on
large, messy, high-dimensional data, at the cost of being slower and less straightforward to interpret. Worth
knowing the name; not needed for a tidy 3-feature dataset like this one.)*

## Part 1: A perceptron for any number of inputs

In the lecture, `perceptron(w1, w2, bias, x1, x2)` hardcoded exactly two inputs. That works fine for two, but
you wouldn't want to write a function with 50 named arguments for a 50-input unit. The standard approach is
to pass the weights as a single **vector**, and the inputs as a single vector too — then the weighted sum is
just `sum(weights * x)`.

**Question 2.** Write a function `sigmoid(x)` that returns `1 / (1 + exp(-x))` (same as the lecture — this
part genuinely doesn't change).

```{r}
# Your function here

```

**Question 3.** Write a function `perceptron(weights, bias, x)`, where `weights` and `x` are numeric vectors
of the *same length* (however long that is), that returns `sigmoid(sum(weights * x) + bias)`.

```{r}
# Your function here

```

**Question 4.** Test your generalized `perceptron()` on the first row of `data`, using
`x = c(data$roundness[1], data$elongation[1], data$symmetry[1])` and a weight vector of your choosing, e.g.
`weights = c(1, 1, 1)` and `bias = 0`. Then try it with only **two** weights and a two-element `x` (say, just
roundness and elongation) — confirm the same function works unchanged for a different number of inputs. That
flexibility is the entire point of writing it this way.

```{r}
# Your code here

```

## Part 2: How wrong is a guess?

**Question 5.** Write a function `rmsd(weights, bias, data, feature_names)`, where `feature_names` is a
character vector naming which columns of `data` to use as inputs (e.g. `c("roundness", "elongation",
"symmetry")`). Inside the function, build each row's input vector from `data[, feature_names]`, run it
through your `perceptron()`, and return the RMSD between `data$category` and the predictions across all rows.

*Hint:* `as.matrix(data[, feature_names])` gives you a matrix where each row is one shape's inputs; you can
loop over rows with a `for` loop, or use `apply()` with `MARGIN = 1` if you're comfortable with it — either
is fine.

```{r}
# Your function here

```

**Question 6.** Try your `rmsd()` with a few different weight vectors and bias values, using all three
features. Can you find a combination with a noticeably lower RMSD than the others?

```{r}
# Your code here

```

## Part 3: The delta rule, generalized

The delta rule from the lecture generalizes the same way. You'll still loop over every row of the dataset,
one example at a time, exactly as before — that part doesn't change. What changes is *within* each row's
update: the lecture wrote one line to update `w1` and a separate line to update `w2`. Here, because `weights`
and `xi` are vectors of the same length, R multiplies them element-by-element, so a single line —
`weights <- weights + lr * delta * xi` — updates the *entire* weight vector at once, however many weights
there are.

$$\Delta \mathbf{w} = \eta \cdot (\text{target} - \text{predicted}) \cdot \text{predicted} \cdot (1 - \text{predicted}) \cdot \mathbf{x}$$

**Question 7.** Fill in the blanks below to complete one training epoch using the generalized delta rule,
on all three features.

```{r}
feature_names <- c("roundness", "elongation", "symmetry")
weights <- c(0.1, 0.1, 0.1)
bias <- 0.1
lr <- 0.5

for (i in 1:nrow(data)) {
  xi <- as.numeric(data[i, feature_names])
  target <- data$category[i]

  pred <- perceptron(weights, bias, xi)
  delta <- (target - pred) * pred * (1 - pred)

  # weights <- weights + lr * ___ * ___
  # bias    <- bias + lr * ___
}

c(weights, bias)
```

**Question 8.** Extend this into a full training loop over 200 epochs, storing the RMSD after each epoch in
a vector called `loss_history`, then plot it. (Same structure as the lecture's training loop — wrap your
Question 7 code in an outer loop over epochs, and calculate the RMSD once at the end of each epoch.)

```{r}
# Your training loop here

```

```{r}
# Your loss_history plot here

```

**Question 9.** Now repeat Questions 7–8 using only **two** features — `c("roundness", "elongation")` — and
compare the final RMSD to the three-feature version. Does adding symmetry actually help here, or does it make
little difference? Relate this back to your answer in Question 1.

```{r}
# Your two-feature training loop here

```

## Part 4: The learning rate

**Question 10.** Re-run your three-feature training loop from Question 8 with a few different learning rates:
something small (`lr <- 0.01`), the original (`lr <- 0.5`), something larger (`lr <- 5`), and something
extreme (`lr <- 50`). Plot `loss_history` for each and compare the shapes of the curves — don't assume in
advance what "too large" will look like; let the plots tell you.

```{r}
# Your code here

```

## Part 5: Back to two dimensions — visualizing a decision boundary

Decision boundaries are easy to draw in 2D but not in 3D, so for this part, go back to using just
`c("roundness", "elongation")` as in Question 9.

**Question 11.** Extend your two-feature training loop to save the weights and bias at epochs 1, 5, 20, and
200, then plot the data with the decision boundary line (`elongation = -(w[1]*roundness + bias) / w[2]`)
overlaid at each checkpoint, using a different colour per checkpoint. (This is the same idea shown in the
lecture — try to reconstruct the checkpointing logic yourself rather than copying it directly.)

```{r}
# Your code here

```

## Part 6: The XOR wall

The generalized `perceptron()` and `rmsd()` you wrote above work for XOR too — it's just another two-feature
dataset.

```{r}
x1 <- c(0, 0, 1, 1)
x2 <- c(0, 1, 0, 1)
category <- c(0, 1, 1, 0)

xor_data <- data.frame(x1, x2, category)
xor_data
```

```{r}
plot(xor_data$x1, xor_data$x2, pch = 19, cex = 3,
     col = ifelse(xor_data$category == 1, "steelblue", "darkorange"),
     xlab = "Input 1", ylab = "Input 2", xlim = c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```

**Question 12.** Train your generalized perceptron on `xor_data` (features `c("x1", "x2")`), using your
Question 8-style training loop. Plot the resulting `loss_history`. Does it ever get close to zero, no matter
how many epochs you run or what learning rate you try?

```{r}
# Your code here

```

**Question 13 (discussion).** Try to find, just by eye, a single straight line on the plot above that puts
both blue points on one side and both orange points on the other. What does this tell you about why the loss
in Question 12 refuses to go down, no matter how you train the network? (Recall the lecture's "What kind of
boundary would work?" slide — in your own words, what *would* it take to separate these four points?)