---
title: "Week 4: Neural Networks 1 -- Solutions"
output:
  html_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

> Posted a few days after the session. Try the exercises yourself first
> -- the value is in getting stuck, not in reading the answer.

## Setup (from the handout)

```{r}
roundness  <- c(0.9, 0.85, 0.8, 0.75, 0.7, 0.65, 0.6, 0.3,
                0.2, 0.15, 0.25, 0.3, 0.35, 0.4, 0.45, 0.7)
elongation <- c(0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.75,
                0.7, 0.8, 0.75, 0.65, 0.6, 0.55, 0.5, 0.2)
symmetry   <- c(0.82, 0.78, 0.88, 0.74, 0.69, 0.64, 0.58, 0.32,
                0.28, 0.22, 0.30, 0.36, 0.42, 0.46, 0.50, 0.60)
category   <- c(1, 1, 1, 1, 1, 1, 1, 1,
                0, 0, 0, 0, 0, 0, 0, 0)
data <- data.frame(roundness, elongation, symmetry, category)

sigmoid <- function(x) {
  1 / (1 + exp(-x))
}
perceptron <- function(weights, bias, x) {
  sigmoid(sum(weights * x) + bias)
}
rmsd <- function(weights, bias, data, feature_names) {
  X <- as.matrix(data[, feature_names])
  predicted <- apply(X, 1, function(x) perceptron(weights, bias, x))
  sqrt(mean((data$category - predicted)^2))
}
```

## Exercise 1

```{r}
plot(data$roundness, data$symmetry, pch = 19,
     col = ifelse(data$category == 1, "steelblue", "darkorange"),
     xlab = "Roundness", ylab = "Symmetry")
```

The two categories separate nearly as cleanly on roundness-vs-symmetry
as they did on roundness-vs-elongation -- Category A (blue) sits in the
high-roundness, high-symmetry corner, Category B (orange) in the
opposite corner. Symmetry looks like it's carrying broadly the same
information as elongation for this dataset, rather than adding an
independent third dimension of separation -- worth keeping in mind for
Exercise 8, where you can check this more directly.

## Exercise 2

There's no single expected answer -- this is exploratory. Weights with a
large positive value on roundness and a large negative value on
elongation (matching how the categories were constructed) reliably push
the output toward 1 for Category-A-like shapes; reversing the signs
pushes toward 0.

## Exercise 3

Already fully worked in the handout. The key result: with
`x1_no_symmetry` (symmetry's value set to 0),
`perceptron(c(2,-2,0), 0, x1_no_symmetry)` and
`perceptron(c(2,-2,5), 0, x1_no_symmetry)` give the **same** output --
changing the symmetry weight from 0 to 5 makes no difference once its
input is 0. Door 1 is shut for that weight this trial, regardless of the
weight's own size.

## Exercise 4

```{r}
x1_two_features <- c(data$roundness[1], data$elongation[1])
perceptron(weights = c(2, -2), bias = 0, x = x1_two_features)
```

`sum(weights * x)` works unchanged because it's computing a dot product,
and a dot product is defined for any pair of equal-length vectors --
there's nothing in the definition of `sum(weights * x)` that hardcodes a
particular length the way writing out `w1*x1 + w2*x2 + w3*x3` by hand
would. Writing the weighted sum with `sum(weights * x)` instead of named
terms is exactly what buys you "works for 2 inputs or 50 inputs, no code
changes" -- the same point Practical 2 made about matrix/vector
operations in general: the operation itself doesn't know or care how
many elements are inside the vectors, only that both vectors match in
length.

## Exercise 5

```{r}
feature_names <- c("roundness", "elongation", "symmetry")
rmsd(c(1, 1, 1), 0, data, feature_names)
rmsd(c(6, -6, 6), 0, data, feature_names)
rmsd(c(8, -8, 8), 0, data, feature_names)
```

`c(6, -6, 6)` gives RMSD around 0.50, versus about 0.57 for `c(1, 1, 1)`
-- confirming the handout's claim. Worth trying `c(8, -8, 8)` too,
though: it actually comes out slightly *worse* (around 0.52), not
better. Scaling all three weights up doesn't help indefinitely -- past
some point, the predictions just saturate harder in whichever direction
they were already leaning, which doesn't fix the points that were on the
wrong side to begin with. The weight combinations that do best all share
the same sign pattern: positive on roundness and symmetry, negative on
elongation -- matching exactly how Category A was constructed (high
roundness/symmetry, low elongation).

The core message is: **a weight's sign should match the direction of the
relationship between its feature and the category**, and the exercise
wants you to *discover* that by trial and error, rather than being told
it upfront.

Walk through why `6, -6, 6` beats `1, 1, 1`: the data was built so
Category A shapes have **high** roundness, **high** symmetry, and
**low** elongation. `6, -6, 6` has a positive weight on roundness, a
negative weight on elongation, and a positive weight on symmetry, so a
shape with high roundness and symmetry but low elongation pushes the
weighted sum strongly positive (toward predicting category A), exactly
matching how the categories were actually constructed. `1, 1, 1` gets
the elongation sign backwards, it treats *higher* elongation as evidence
*for* category A, when in the real data it's evidence against it. That
mismatch on just one weight is enough to visibly hurt the fit.

## Exercise 6

Already fully worked in the handout. `w_new_a` and `w_new_b` differ only
in the **third** entry (the symmetry weight) -- it moves in `w_new_a`
(where symmetry's real value feeds in) and stays frozen in `w_new_b`
(where symmetry was zeroed out for that trial). The first two entries
(roundness, elongation weights) are identical in both, since neither
trial changed those inputs. Zeroing out a *different* feature (e.g.
roundness instead of symmetry) freezes a different weight -- whichever
one matches the zeroed-out input, confirming this isn't about symmetry
specifically, it's about door 1 being shut for whichever weight's own
input was zero.

## Exercise 7

Already fully worked in the handout. `delta_unsure` (starting weights
all zero, `predicted = 0.5`) comes out close to 5 times larger than
`delta_confident` (starting weights `c(2,-2,0)`, `predicted` around
0.83) -- same door-2 pattern as the lecture, now with three inputs
feeding the same sigmoid. A third, in-between starting-weight choice
should give a `delta` somewhere between the two extremes -- roughly
speaking, the closer `predicted` lands to 0.5, the closer `delta` will
be to the unsure end; the closer to 0 or 1, the closer to the confident
end.

## Exercise 8

```{r}
feature_names <- c("roundness", "elongation", "symmetry")
weights <- c(0.1, 0.1, 0.1); bias <- 0.1; lr <- 0.5; n_epochs <- 200
loss_history <- numeric(n_epochs)
for (epoch in 1:n_epochs) {
  for (i in 1:nrow(data)) {
    xi <- as.numeric(data[i, feature_names])
    target <- data$category[i]
    pred <- perceptron(weights, bias, xi)
    delta <- (target - pred) * pred * (1 - pred)
    weights <- weights + lr * delta * xi
    bias <- bias + lr * delta
  }
  loss_history[epoch] <- rmsd(weights, bias, data, feature_names)
}
loss_history[n_epochs]   # about 0.357

feature_names2 <- c("roundness", "elongation")
weights2 <- c(0.1, 0.1); bias2 <- 0.1
loss_history2 <- numeric(n_epochs)
for (epoch in 1:n_epochs) {
  for (i in 1:nrow(data)) {
    xi <- as.numeric(data[i, feature_names2])
    target <- data$category[i]
    pred <- perceptron(weights2, bias2, xi)
    delta <- (target - pred) * pred * (1 - pred)
    weights2 <- weights2 + lr * delta * xi
    bias2 <- bias2 + lr * delta
  }
  loss_history2[epoch] <- rmsd(weights2, bias2, data, feature_names2)
}
loss_history2[n_epochs]   # about 0.361
```

The two-feature and three-feature final RMSDs come out almost identical
(about 0.361 vs 0.357) -- adding symmetry makes only a marginal
difference here. That matches your Exercise 1 observation: symmetry
separates the categories in almost the same way elongation already does,
so it isn't contributing much genuinely new information for the network
to use. Neither version gets especially close to 0, which is expected --
this dataset has some inherent noise built into it (not every point sits
exactly where its category "should" place it), and no perceptron,
however well trained, can perfectly fit noise that isn't really there to
be fit.

## Exercise 8b (optional)

```{r}
# (run right after the three-feature Exercise 8 training)
feature_names <- c("roundness", "elongation", "symmetry")
X <- as.matrix(data[, feature_names])
predicted_prob <- apply(X, 1, function(x) perceptron(weights, bias, x))
predicted_category <- ifelse(predicted_prob >= 0.5, 1, 0)

data.frame(prob = round(predicted_prob, 2),
           predicted = predicted_category,
           actual = data$category)
```

The trained network classifies 14 of the 16 shapes "correctly" and gets
exactly two "wrong": **row 8** and **row 16**. But look those two up in
the data and find them in your Exercise 1 scatter plots: row 8 is
labeled Category A yet sits squarely among the Category B points (low
roundness, high elongation, low symmetry), and row 16 is the mirror
image. These are the dataset's two noisy points as their labels don't
match where they actually sit. The network isn't confused about them at
all: it's *confidently* on the "wrong" side for both (probability around
0.07 for row 8 and 0.78 for row 16), because it learned the overall
pattern and those two labels contradict the pattern.

This is also exactly why the Exercise 8 RMSD levels off well above 0
rather than approaching it: those two shapes contribute a large,
permanent error that no amount of training can remove as a single
straight-line boundary can't chase two points sitting deep in the other
category's territory without misclassifying several well-behaved points
instead. That inability turns out to be a feature, not a bug: refusing
to contort itself around noisy labels is what keeps the model's boundary
sensible.

## Exercise 9

```{r}
run_training <- function(lr_value, n_epochs = 200) {
  weights <- c(0.1, 0.1, 0.1); bias <- 0.1
  loss_history <- numeric(n_epochs)
  for (epoch in 1:n_epochs) {
    for (i in 1:nrow(data)) {
      xi <- as.numeric(data[i, feature_names])
      target <- data$category[i]
      pred <- perceptron(weights, bias, xi)
      delta <- (target - pred) * pred * (1 - pred)
      weights <- weights + lr_value * delta * xi
      bias <- bias + lr_value * delta
    }
    loss_history[epoch] <- rmsd(weights, bias, data, feature_names)
  }
  loss_history
}

plot(run_training(0.01), type = "l", main = "lr = 0.01", ylim = c(0.3, 0.8))
plot(run_training(0.5),  type = "l", main = "lr = 0.5",  ylim = c(0.3, 0.8))
plot(run_training(5),    type = "l", main = "lr = 5",    ylim = c(0.3, 0.8))
plot(run_training(50),   type = "l", main = "lr = 50",   ylim = c(0.3, 0.8))
```

`lr = 0.01` descends very slowly and clearly hasn't finished descending
by epoch 200; it's still on its way down when the plot ends. `lr = 50`
descends smoothly and settles into a low, stable RMSD well before 200
epochs. `lr = 5` and `lr = 50` both reach a similar final RMSD to
`lr = 0.5` , but the path there is visibly rougher; the loss curve jumps
around noticeably before settling, especially at `lr = 50` , rather than
descending smoothly.

What `eta` is actually controlling: the size of the jump the weights
make across the error landscape after each trial, the same landscape you
visualized as a cost surface in the earlier lectures. A larger `eta`
means bigger jumps across that surface per step.

That can cut both ways. A bigger jump **can** be an advantage on a
bumpier landscape than this one; if the surface has more than one
valley, a small step size can get trapped in whichever valley it happens
to land near first, while a larger step is more likely to jump clean
over a shallow ridge and land somewhere better. (For this particular
dataset, all four learning rates end up in roughly the same place, so
this benefit isn't something you can actually see here, but it's worth
knowing as a real, general possibility on other, bumpier landscapes.)

The downside is more visible in what you just plotted. A jump that's too
large can overshoot the bottom of the valley entirely, landing further
up the **other** side rather than settling in, which is exactly the
bouncing you can see in the `lr = 50` curve, each step correcting too
hard and overcorrecting again. In the worst case, an oversized jump
early in training can push a unit's weights so far that its prediction
saturates all the way to the wrong extreme (predicted very close to 0 or
1 when it should be the opposite). Once that happens, training can get
stuck rather than just noisy: recall from the lecture that
`predicted * (1 - predicted)` vanishes as predicted approaches 0 or 1,
so a badly saturated, wrong prediction gives almost no correction signal
to fix itself with. A learning rate large enough to cause that isn't
just inefficient, it can genuinely break training, which is the real
reason "too large" is a danger worth taking seriously, not just a matter
of a slightly bumpier-looking loss curve.

## Exercise 10

```{r}
feature_names <- c("roundness", "elongation")
weights <- c(0.1, 0.1); bias <- 0.1
checkpoints <- c(1, 5, 20, 50, 200)
saved <- list()
for (epoch in 1:200) {
  for (i in 1:nrow(data)) {
    xi <- as.numeric(data[i, feature_names])
    target <- data$category[i]
    pred <- perceptron(weights, bias, xi)
    delta <- (target - pred) * pred * (1 - pred)
    weights <- weights + 0.5 * delta * xi
    bias <- bias + 0.5 * delta
  }
  if (epoch %in% checkpoints) saved[[as.character(epoch)]] <- c(weights, bias)
}

plot(data$roundness, data$elongation, pch = 19,
     col = ifelse(data$category == 1, "steelblue", "darkorange"),
     xlab = "Roundness", ylab = "Elongation")
x_seq <- seq(0, 1, length.out = 100)
cols <- c("gray85", "gray65", "gray45", "gray25", "black")
for (k in seq_along(checkpoints)) {
  w <- saved[[as.character(checkpoints[k])]]
  lines(x_seq, -(w[1] * x_seq + w[3]) / w[2], col = cols[k], lwd = 2)
}
legend("bottomleft", legend = paste("epoch", checkpoints), col = cols, lwd = 2)
```

The epoch-50 line sits much closer to the epoch-200 line than to the
epoch-20 line -- most of the boundary's total movement has already
happened by epoch 50, and training after that mostly fine-tunes rather
than making large jumps. This matches a pattern that shows up constantly
in gradient-based training generally: early updates, when the model is
furthest from a good fit, tend to be the largest; later updates get
progressively smaller as the model settles in, even with a constant
learning rate -- because `delta` itself shrinks as `predicted` gets
closer to `target` for most points.

## Exercise 11

```{r}
weights <- c(0.1, 0.1); bias <- 0.1
n_epochs <- 2000; lr <- 5
x1 <- c(0, 0, 1, 1); x2 <- c(0, 1, 0, 1); category_xor <- c(0, 1, 1, 0)
xor_data <- data.frame(x1, x2, category = category_xor)

for (epoch in 1:n_epochs) {
  for (i in 1:nrow(xor_data)) {
    xi <- c(xor_data$x1[i], xor_data$x2[i])
    target <- xor_data$category[i]
    pred <- perceptron(weights, bias, xi)
    delta <- (target - pred) * pred * (1 - pred)
    weights <- weights + lr * delta * xi
    bias <- bias + lr * delta
  }
}
# same rmsd() reuse as the handout's training loop -- xor_data has a category
# column, so the function works unchanged
rmsd(weights, bias, xor_data, c("x1", "x2"))
```

No -- however you set the learning rate (tried here: much larger,
`lr = 5`) or however many epochs you allow (tried here: 2000, ten times
the original), the final RMSD stays close to 0.5, essentially unchanged
from a short run at the default settings. This is the clearest possible
confirmation that XOR isn't a "hasn't trained long enough yet" problem
-- more training time and different learning rates don't touch it,
because the limitation is structural, not a matter of search effort.

## Exercise 12 (discussion)

There's no line through the four XOR points that separates the two blue
points from the two orange points -- every candidate line either has a
blue point on the wrong side, an orange point on the wrong side, or
both. This is the direct, visual reason the loss never approaches zero:
the delta rule is very good at finding the best possible straight-line
boundary, but for XOR, even the best straight line still gets some
points wrong, no matter how it's rotated or shifted. As the lecture's "A
puzzle" slide previewed, actually solving this needs more than a
straight line. Specifically, the second-layer, multi-unit approach
covered in next week's lecture, where multiple simple boundaries get
combined into something that can carve out the diagonal pattern XOR
requires.
