---
title: "Week 3: Resampling & Monte Carlo Methods"
subtitle: "How Randomness can help us Understand the World"
output:
  html_document:
    toc: true
format:
  html:
    toc: true
    toc-expand: 2
---

## Install packages

This exercise requires the packages `car`, `tidyr` and `afex`. You need to install the packages once on your computer. To do this, you please run the following code, once on the console (copy & paste)

`install.packages(c("afex", "car", "tidyr"))`

Before you start the exercises, please always load the required packages by running the following code:

```{r, message=FALSE}
library("afex")
library("car")
library("tidyverse")
```

##  Randomness
Randomness is a fundamental concept in statistics and data science. It plays a crucial role in many areas, including simulation, sampling, and hypothesis testing.

### Exercise 1: Random Numbers

You can generate 10 random numbers from a normal distribution in R using the `rnorm()` function. The result is stored in the variable `v`.

```{r, message=FALSE}
v = rnorm(5, mean = 0, sd = 1)
print(v)
```

Generate 5 random numbers with a mean of 5 and a standard deviation of 2 and calculate the mean and standard deviation of the generated numbers using the function `mean()` and `sd()`.

```{r, message=FALSE}
#your code here

```

### Exercise 2: Random Numbers
**(a)** Re-run the code multiple times. Is it exactly the same? How close is the mean and standard deviation to the specified values of 5 and 2?



**(b)** What happens if you change the sample size to e.g. 100, or 10,000? How does it change the mean and standard deviation? Or how robust are the mean and standard deviation of the distribution if you make new simulations with different sample sizes?



## Resampling

The dataset below contains 10 individuals' self-esteem score on three time points during a specific diet to determine whether their self-esteem improved.

We load the data `selfesteem`:


```{r, message=FALSE}
dat_text = "
id       t1       t2       t3
1 400.5027 518.2286 710.7831
2 255.8124 691.2915 630.8434
3 324.4241 444.3434 977.8410
4 341.9538 471.1696 834.7124
5 287.1243 390.8429 645.7287
6 204.5868 534.0549 665.3224
7 352.5992 558.0695 684.0157
8 317.9425 437.0234 781.8623
9 350.7964 439.9808 847.1229
10 304.3798 448.9376 858.1100
"
selfesteem = read.table(text=dat_text, header = TRUE)
```

The data look like this:
```{r, message=FALSE}
plt  = Boxplot(selfesteem[c("t1", "t2", "t3")], main = "Boxplot of Self-Esteem Scores", ylab = "Self-Esteem Score")
```

### Exercise 3: Median

The median for `t1`is calculated like this:
```{r, message=FALSE}
median(selfesteem$t1)
```

Use the function to calculate the median and standard deviation of the self-esteem scores at time point 1 (`t2`, `t3`).

```{r, message=FALSE}
#your code here

```


### Exercise 4: Resampling

We want to calculate the confidence intervals for the medians at all time points. To do this, we will use bootstrapping to estimate the sampling distribution of the median.

```{r, message=FALSE}
sampling_dist = c()
for (i in 1:3) {
  resampled_dat = sample(selfesteem$t1, size=10, replace = TRUE)
  median_resampled = median(resampled_dat)
  sampling_dist = c(sampling_dist, median_resampled)
}
```

The code above produces only very few resamples of the median. Change the code that you produce 1000 resample.

```{r, message=FALSE}
#your code here

```

You can visualize the created sampling distribution `sampling_dist` using a histogram.

```{r, message=FALSE}
hist(sampling_dist, main = "Sampling Distribution of the Median", xlab = "Median", nclass = 10, col = "lightblue")
```

### Exercise 5: Resampled Data

**(a)**: Run the resampling multiple time and look at the histogram. Also increade and decrease  the number of resamples (e.g. 100 and 10000). Run it also multiple times and see how the resulting sampling behaves.

How does the number of resamples affect the shape of the sampling distribution? What is the effect of the number of resamples when you run it multiple times?



**(b)**: On what does the reproducibility of the sampling distribution depend?



### Exercise 6: Sampling Distribution

**(a)**: If you have the sampling distribution of the median for `t1`, you can calculate the confidence intervals for the median. Calculate 95% bootstrapped confidence intervals, which are based on the 2.5th and 97.5th percentiles (quantiles) of the sampling distribution (see slides).

```{r, message=FALSE}
#your code here

```

**(b)**: Calculate the 90% and 99% confidence intervals for the medians (`t1`).

```{r, message=FALSE}
#your code here

```

### Exercise 7: Confidence Intervals

Calculate the confidence intervals for the medians at all time points. copy, paste & modify the code:

**(a)**: Bootstrapped confidence interval for `t2`:

```{r, message=FALSE}
#your code here

```

**(b)**: Bootstrapped confidence interval for `t3`:

```{r, message=FALSE}
#your code here

```


### Exercise 8: Example Data

We examine the mathematica anxiety of students from different faculties. Low values indicate low anxiety, high values indicate high anxiety. The data contain the following columns:

- `subject_id`: The ID of the subject
- `faculty`: The faculty the subject belongs to
- `anxiety`: The level of anxiety

We download the data into the data frame called `math_anxiety_dat`:

```{r, message=FALSE}
url = "https://raw.githubusercontent.com/lindemann09/block-41p-ds-2026/refs/heads/main/data/math_anxiety.csv"
math_anxiety_dat =read.csv(url)
head(math_anxiety_dat)
```

```{r, message=FALSE}
plt  = Boxplot(math_anxiety_dat, main = "Math Anxiety Scores", ylab = "Math Anxiety Score")
```

Calculate the means and standard deviations for psychology, sport science and medicine:

```{r, message=FALSE}
#your code here

```

### ANOVA is R

This is the code for a one-way ANOVA to test for differences between the three groups:

```{r, message=FALSE}
# create dataframe for anova (data have to be in long format and an id column is needed)
anova_dat = pivot_longer(math_anxiety_dat,
          cols = everything(),
          names_to = "faculty",
          values_to = "anxiety")
anova_dat$id = 1:nrow(anova_dat)
res = aov_ez(id ="id", dv = "anxiety", between = "faculty", data = anova_dat)
print(res)
```

The effect is not significant.

### Exercise 9: Simulate ANOVA Data

Let's make a simulation-based power analysis to estimate the power of the ANOVA with 30 participants per group.

We simulate data and use the means and standard deviations from the real data (round to two decimal places). Please complete the code (`....`) to simulate a data set.

```{r, eval=FALSE}
anxiety_simulated = data.frame(
  psychology    = rnorm(...., ...., ....),
  sport_science = rnorm(...., ...., ....),
  medicine      = rnorm(...., ...., ....))
```

```{r, message=FALSE}
#your code here

```

### Exercise 10: Simulation ANOVAs Outcomes

Simulate 1000 data sets and calculate the proportion of significant results (p < 0.05) to estimate the power of the ANOVA with 30 participants per group.

Please complete the code (`....`) correctly:

```{r, eval=FALSE, message=FALSE}
results = c()
for (i in 1:......) {
  # 1. simulation
  anxiety_simulated = ......

  # 2. perform ANOVA
  anova_dat = pivot_longer(anxiety_simulated,
                           cols = everything(),
                           names_to = "faculty",
                           values_to = "anxiety")
  anova_dat$id = 1:nrow(anova_dat)
  res = aov_ez(id ="id", dv = "anxiety", between = "faculty",
                data = anova_dat)
  # get the p-value from the results
  p_value = res$anova_table$`Pr(>F)`

   # 3. check significance
  is_sig =  .......

  # store the result
  results = c(results, is_sig)
}
```


```{r, message=FALSE}
#your code here

```


### Exercise 11: Simulation-based Power Analysis

**(a)**: What the bootstrap power of the ANOVA with 30 participants per group?

```{r, message=FALSE}
#your code here

```

**(b)**: Interpret the result. What does it mean?



### Exercise 12: A-priori Power Analysis

A-priori power analysis asks the question: What sample size do we need to reach a power of 80%. That is, how many subjects are needed that we find an effect with a 80% probability, if there is an effect in the real world?

**(a)**: We can increase the power by increasing the sample size. What is the power if we have more participants per group? Change sample size of the simulated data and see how the power changes.

```{r, message=FALSE}
#your code here

```

**(b)**: How many participants per group do we need to reach a power of 80%?


