---
title: "Week 3: Parameter Estimation"
subtitle:  "Basic Methods & Maximum Likelihood"
---

## Before you start

This tutorial uses the same psychometric-function dataset from the lecture. You'll rebuild the grid search
and `optim()` fits yourself, then learn how a dataset like this could be simulated with known parameters (so
you can check whether your methods recover the truth), and finally compare today's Weibull model against
the linear model you already know from statistics.

You don't need to understand every line of R syntax perfectly — focus on understanding *what each step is
doing conceptually*. If something errors, re-check you've run all the chunks above it in order.

## Part 0: Setup and data

Run this chunk first — this is the same dataset from the lecture, presented as if it were real data (there's
no "true" underlying parameter values to reveal here, just like with a real participant).

```{r}
intensity <- 1:9
n_trials <- rep(40, 9)
n_yes <- c(0, 4, 8, 16, 25, 28, 37, 39, 40)
prop_yes <- n_yes / n_trials

data <- data.frame(intensity, n_yes, n_trials, prop_yes)
data
```

```{r}
psychometric_model <- function(alpha, beta, x) {
  y <- rep(0, length(x))
  valid <- x >= 0
  y[valid] <- 1 - exp(-(x[valid] / alpha)^beta)
  y
}

plot(data$intensity, data$prop_yes, pch = 19,
     xlab = "Stimulus intensity", ylab = "Proportion 'yes' responses",
     ylim = c(0, 1))
```

**Question 1.** Just by looking at the table and plot above, roughly guess values of `alpha` (scale — where
the curve is around 63%) and `beta` (shape — how sharply it rises). Write your guesses down; you'll compare
them to a proper fit shortly.

## Part 1: Grid search by hand

```{r}
rmsd <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$intensity)
  sqrt(mean((data$prop_yes - predicted)^2))
}
```

**Question 2.** Use the `rmsd()` function to compare your eyeballed guess from Question 1 against two other
guesses of your choice. Which has the lowest RMSD?

```{r}
# Try it here: rmsd(c(alpha_guess, beta_guess), data)

```

Now do a full grid search:

```{r}
alpha_vals <- seq(1, 9, length.out = 100)
beta_vals <- seq(0.1, 6, length.out = 100)
grid <- expand.grid(alpha = alpha_vals, beta = beta_vals)
grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))

best <- grid[which.min(grid$rmsd), ]
best
```

```{r}
library(lattice)
levelplot(rmsd ~ alpha * beta, data = grid, xlab = "alpha (scale)", ylab = "beta (shape)")
```

**Question 3.** How close was your eyeballed guess to the grid search result? What does the dark "valley"
region in the plot represent?

## Part 2: Automatic search with `optim()`

```{r}
fit_rmsd <- optim(par = c(alpha = 4, beta = 1), fn = rmsd, data = data)
fit_rmsd$par
```

**Question 4.** Change the starting guess (the `par` argument) to something far off, e.g. `c(alpha = 1,
beta = 5.5)`, and re-run. Does `optim()` still find (roughly) the same answer? Try a really extreme starting
guess — can you get it to fail or get stuck? (Recall from the lecture: `optim()`'s default method is
Nelder-Mead simplex, which can in principle get stuck in a local minimum — though for this particular
two-parameter problem, the RMSD surface has only one clean valley, so it's unlikely to happen here.)

```{r}
# Try alternative starting values here

```

**Question 5.** Compare `fit_rmsd$par` to your Question 1 eyeball guess. Which do you trust more, and why?
(Notice there's no "true" value here to check against, since this is real-looking data, not simulated data —
that's exactly why systematic fitting matters more than eyeballing, not less: in real research you usually
don't have a ground truth to compare to.)

## Part 3: Maximum likelihood estimation

```{r}
nll <- function(params, data) {
  alpha <- params[1]; beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$intensity)
  predicted <- pmin(pmax(predicted, 1e-6), 1 - 1e-6)
  -sum(dbinom(data$n_yes, data$n_trials, predicted, log = TRUE))
}

fit_mle <- optim(par = c(alpha = 4, beta = 1), fn = nll, data = data)
fit_mle$par
```

**Question 6.** Compare `fit_mle$par` to `fit_rmsd$par`. Are they similar?

**Question 7 (discussion).** RMSD asks "which parameters minimize the average squared prediction error?"
MLE asks "which parameters make the observed data most probable?" In your own words, why might these two
questions often — but not always — lead to similar answers?

*Hint:* try comparing what happens at an intensity where `prop_yes` is close to 0.5 versus one where it's
close to 0 or 1 (check the `data` table from Part 0 for examples of both). Does a "miss" of the same size —
say, the model predicting 2 fewer "yes" responses than observed — feel like it should count the same in both
cases? RMSD treats every miss the same regardless of where it happens; does likelihood?

## Part 4: Simulating your own data

So far you've worked with a fixed dataset. In real research, it's common to simulate data from known
parameters first, to check that your fitting methods actually recover the truth before trusting them on real
data. Here's how, using the same `rbinom()` function from Practical 2, together with the Weibull formula from
`psychometric_model()` to generate the underlying probabilities.

```{r}
set.seed(99)
intensity_sim <- 1:9
true_alpha <- 4
true_beta <- 2
p_true <- psychometric_model(true_alpha, true_beta, intensity_sim)

n_trials_sim <- rep(40, 9)
n_yes_sim <- rbinom(length(intensity_sim), n_trials_sim, p_true)
prop_yes_sim <- n_yes_sim / n_trials_sim

data_sim <- data.frame(intensity = intensity_sim, n_yes = n_yes_sim,
                        n_trials = n_trials_sim, prop_yes = prop_yes_sim)
data_sim
```

```{r}
plot(data_sim$intensity, data_sim$prop_yes, pch = 19,
     xlab = "Stimulus intensity", ylab = "Proportion 'yes' responses", ylim = c(0, 1))
```

This is the general pattern for simulating any dataset: pick true parameter values, feed them into your model
to get predicted probabilities, then use `rbinom()` to add realistic trial-to-trial noise on top of those
predictions.

**Question 8.** Using either `rmsd()` + `optim()`, or `nll()` + `optim()` (your choice), fit the
psychometric model to `data_sim`. Report your estimated `alpha` and `beta`.

```{r}
# Your fit here

```

**Question 9.** Compare your fitted `alpha` and `beta` to `true_alpha` and `true_beta` in the chunk above.
How close is the recovered estimate to the parameters that actually generated the data?

**Question 10.** Plot the fitted curve on top of `data_sim`. Here's a scaffold to get you started — replace
the two blanks with your fitted `alpha` and `beta` from Question 8 (e.g. `fit_sim$par[1]` and
`fit_sim$par[2]`, assuming you called your fitted object `fit_sim`):

```{r}
plot(data_sim$intensity, data_sim$prop_yes, pch = 19,
     xlab = "Stimulus intensity", ylab = "Proportion 'yes' responses", ylim = c(0, 1))

x_seq <- seq(0, 9, length.out = 200)
# lines(x_seq, psychometric_model(___, ___, x_seq), col = "blue")
```

If the line doesn't appear, double check: (1) the `lines()` call above is uncommented (no `#` in front), (2)
you've replaced both blanks with actual numbers or variable references, and (3) you ran the fit in Question
8 in the same session, so the object you're referencing still exists.

**Question 11 (interpretation).** In plain psychological language (not equations), what do this simulated
participant's `alpha` and `beta` tell you about their sensitivity, compared to the participant from Part
0–3? Does this participant have a higher or lower scale (sensitivity), a sharper or shakier transition, or
some combination?

**Question 12 (extension).** Since you can see (and edit) the simulation code above, try changing
`true_alpha` and/or `true_beta` to values of your own choosing — pick something quite different from 4 and 2
(e.g. a much lower `alpha`, or a very shallow or very steep `beta`) — then re-run that chunk and repeat
Questions 8–10 on your new dataset. Does your fitting procedure still recover your chosen values accurately?
Are there any combinations where recovery seems to work noticeably less well, and if so, what do you think
might be going on (hint: think back to the lecture's discussion of what makes a landscape easy or hard to
search)?

## Part 5: Does the shape of the model matter?

So far we've assumed detection follows a Weibull curve. But you already know a simpler model from your
statistics background: a straight line, $y = a + b \cdot x$. Let's see how well it does on the **same data
from Part 0–3** (the `data` object), compared to the Weibull model you already fit (`fit_rmsd`).

This part is more independent than the previous ones — you're writing the functions from scratch, based on
the pattern of `psychometric_model()` and `rmsd()` you've already used, rather than adapting code that's
given to you.

**Question 13.** Write a function `linear_model(a, b, x)` that returns `a + b * x`.

```{r}
# Your function here

```

**Question 14.** Write a function `rmsd_linear(params, data)`, following the same structure as `rmsd()`, but
using `linear_model()` instead of `psychometric_model()`.

```{r}
# Your function here

```

**Question 15.** Fit the linear model to `data` using `optim()` (try a starting guess like
`c(a = 0, b = 0.1)`). Report the fitted `a` and `b`, and the resulting RMSD value (`$value`).

```{r}
# Your fit here

```

**Question 16.** Compare the RMSD of your linear fit to the RMSD of the Weibull fit (`fit_rmsd$value`).
Which model achieves a lower RMSD on this data?

**Question 17.** Plot `data` again, and overlay *both* fitted curves (Weibull in one color, linear in
another) using `lines()`. Visually, where do the two models agree, and where do they most clearly diverge?

```{r}
# Your plot here

```

**Question 18 (discussion).** A lower RMSD isn't automatically "the better model" — a model with more
flexibility can often fit any given dataset more closely, whether or not it's a good theory of the
underlying process. Just based on what you can see in your plot from Question 17: `prop_yes` is a
*proportion*, so it can only ever be between 0 and 1. Does the linear model's fitted line respect that
constraint across the full intensity range shown, or does it predict something impossible somewhere?
(Hint: check what the linear model predicts at intensity = 1 and at intensity = 9.) We'll cover how to
formally judge whether extra fit flexibility is "worth it" in the Model Comparison sessions later in the
course — for now, just describe what you notice.
