---
title: "Practical 1: Introduction 1"
subtitle: "The R Basics"
output:
  html_document:
    toc: true
---

## Before you start

This session is entirely exercises based on our
[R Basics for Data Science](https://lindemann09.github.io/block-41p-ds-2026/r_basics/) tutorial, which you
should read (or at least skim) before this session. Nothing here repeats his explanations; each part below
just gives a one-line reminder of what it's practicing, then gets straight to exercises. The one exception is
Part 3 ("Numerical gotchas"), which covers a few special values our manual doesn't — you'll see this
same idea again later in the course.

If you get stuck, the fix is almost always: go back and re-read the matching section of our manual, or
use `?functionname` to check what a function actually expects.

-------------------------------------------

## Part 1: Variables and basic operations

*(Reminder: `<-` assigns a value to a name; R evaluates arithmetic the way a calculator would.)*

### Exercise 1

 A study collects three reaction times, in milliseconds: 452, 618, and 390. Store each in its
own variable, then compute their average by hand (adding and dividing) — don't use `mean()` yet, that comes
later.

```{r}
# Your code here

```

### Exercise 2

 A participant's score starts at 0. Over three trials they gain 4, then lose 1, then gain 2
points. Using the "update a variable by reassigning it" pattern from the manual (`x <- x + 1`), track their
score through all three trials, printing it after each one.

```{r}
# Your code here

```

### Exercise 3

 If `r` is the radius of a circle, the area is $\pi r^2$. Compute the area of a circle with
radius 7.5 (R has `pi` built in).

```{r}
# Your code here

```

----------------------------------

## Part 2: Data types

*(Reminder: numbers, character strings, and logicals (`TRUE`/`FALSE`) are different types; R will sometimes
convert between them, which can cause quiet, confusing bugs.)*

### Exercise 4

 Store the text `"25"` in a variable called `age_text`. Try multiplying it by 2. What happens,
and why? Now fix it using the appropriate `as.` function from the manual, and try again.

```{r}
# Your code here

```

### Exercise 5

 Without running any code, predict what each of these will print. Then run them to check.

```r
5 > 3
5 == "5"
!(5 > 3)
"cat" == "dog"
```

### Exercise 6

 A participant's response was recorded as `TRUE` (they pressed the correct key) or `FALSE`
(they didn't). Write one line of code that converts a logical value `responded_correctly <- TRUE` into the
text `"correct"` or `"incorrect"`. *(Hint: you don't need an if-statement for this — search for `ifelse()`
in R's help, or just use `paste()` creatively. If-statements themselves are next session.)*

```{r}
# Your code here

```

----------------------------------

## Part 3: Numerical gotchas — when R goes quiet instead of throwing an error

You've already met `NA` ("not available") in our manual. R has a few other special values that behave
similarly — and unlike a normal error, which stops your code and shows you a clear red message, these values
often just slot silently into your result and keep going. That's what makes them genuinely dangerous: your
code doesn't crash, it just quietly starts computing the wrong thing.

### Exercise 7

 Predict, then check, what each of these produces:

```r
1 / 0
-1 / 0
0 / 0
log(0)
log(-1)
```

You should see two new special values: `Inf` (and `-Inf`), meaning "a number too large to represent, in that
direction" — not an error, a genuine value R will happily keep computing with — and `NaN`, meaning "not a
number," R's way of saying "this calculation has no sensible numeric answer at all."

### Exercise 8

 R gives you three functions to check for these values: `is.na()`, `is.nan()`, and
`is.finite()`. Run all three of them on each of `NA`, `NaN`, `Inf`, and an ordinary number like `5`, and fill
in which combinations return `TRUE`. One result should genuinely surprise you — `is.na(NaN)`. Look up
`?is.na` to find out why R treats `NaN` as a special case of `NA`, rather than as something completely
separate.

```{r}
# Your code here

```

### Exercise 9 (discussion)

 Run this:

```{r}
values <- c(3, 5, NaN, 2, 8)
sum(values)
mean(values)
```

Even though 4 of the 5 numbers are completely ordinary, the result is `NaN` — one bad value was enough to
"poison" the entire calculation. This is a preview of something you'll run into for real later in the
course: a single `log(0)` buried inside a much larger calculation (like a model's total fit to a whole
dataset) can silently turn the *entire* result into `-Inf` or `NaN`, not just the one data point that
actually caused it. Keep this pattern in mind — you'll meet the full version of this problem, and the fix
for it, in the Supervised Learning practical.

----------------------------------

## Part 4: Calling functions

*(Reminder: functions are called with `name(argument1, argument2, ...)`; use `?functionname` to see what
arguments a function takes and which have defaults.)*

### Exercise 10

 Look up the help documentation for `round()`. Use it to round 3.14159 to 2 decimal places.

```{r}
# Your code here

```

### Exercise 11

 Generate 8 random numbers from a normal distribution with mean 100 and standard deviation
15 (same distribution mentioned in the manual for IQ scores).

```{r}
# Your code here

```

### Exercise 12

 Nest three function calls together in one line: take the square root of the natural log of
50.

```{r}
# Your code here

```

----------------------------------

## Part 5: Vectors

*(Reminder: `c()` builds a vector; operations on a vector apply to every element at once; `[]` extracts
elements by position, by exclusion, or by a logical condition.)*

### Exercise 13

 Store these five reaction times (ms) in a vector: 452, 618, 390, 705, 511.

```{r}
# Your code here

```

### Exercise 14

 Using your Exercise 13 vector: compute the mean and the maximum with built-in functions
(not by hand this time), then subtract the mean from every value in one line (this is called
"mean-centering," and you'll meet it again later in the course).

```{r}
# Your code here

```

### Exercise 15

 From the same vector, extract: (a) the 2nd and 4th elements; (b) every value greater than
500; (c) all values *except* the maximum. *(Hint for (c): combine `which.max()` with negative indexing.)*

```{r}
# Your code here

```

### Exercise 16

 Create a named vector storing the ages of three lab members: Alex (28), Priya (34), and Sam
(25). Retrieve Priya's age using her name, not her position.

```{r}
# Your code here

```

### Exercise 17

 Generate a sequence of 10 evenly-spaced numbers between 0 and 1 (you've seen the function
for this both in the manual and already in the course's own lecture material).

```{r}
# Your code here

```

### Exercise 18 (discussion)

 Exercise 14 asked you to subtract the mean from *every* value in your vector,
in a single line, with no loop. Why does this work? What is R actually doing, one element at a time, behind
that one line of code? Keep this Exercise in mind — next session's whole final section is about exactly this
idea.

