---
title: "Block 4.1 Practical: Data Science "
subtitle: "Erasmus Univeristy Rotterdam"
author:
  - name: "Oliver Lindemann"
  - name: "Yaniv Morgenstern"

format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
---

This manual website contains all instructions and the important study materials for the course "Data Science". Further information about this course are found on [Canvas](https://canvas.eur.nl/courses//55946).

## Schedule

```{r}
#| echo: false
week <- function(x) {
  w = as.Date('1.9.2026',format='%d.%m.%Y') + 7*(x-1)
  txt = format(w, format = "%d/%m")
  return(txt)
}

# linked materials
visible_practicals = 2
visible_lectures = 0

practical_title = c(
  "Getting Started with R I",
  "Getting Started with R II",
  "Parameter Estimation",
  "Neural Networks 1",
  "Neural Networks 2",
  "Cognitive Models in Psychology",
  "Cognitive Models I",
  "Cognitive Models II",
  "Cognitive Models III")

lecture_tile= c(
  "Introduction to Data Science",
  "Parameter Estimation",
  "Neural Networks 1",
  "Neural Networks 2",
  "Cognitive Models in Psychology",
  "Random walk methods",
  "Reaction Time Models",
  "Q & A Lecture")

## week link
practical <- function(week) {
  if (week <= visible_practicals) {
    return(paste0("[", practical_title[week], "](practicals/data_science_week", week, ".Rmd)"))
  } else {
    return(practical_title[week])
  }
}

lecture <- function(week) {
  if (week <= visible_lectures) {
    return(paste0("[", lecture_tile[week], "](lectures/lecture", week, ".pdf)"))
  } else {
    return(lecture_tile[week])
  }
}


```

| Week        | Topic | Tutorial   (11:00) | Lecture (13:00) |
| ----------- | ----- | ------------------ | --------------- |
| `r week(1)` | 1     | `r practical(1)`   | `r lecture(1)`  |
| `r week(2)` | 2     | `r practical(2)`   | `r lecture(2)`  |
| `r week(3)` | 3     | `r practical(3)`   | `r lecture(3)`  |
| `r week(4)` | 4     | `r practical(4)`   | `r lecture(4)`  |
| `r week(5)` | 5     | `r practical(5)`   | `r lecture(5)`  |
| `r week(6)` | 6     | `r practical(6)`   | `r lecture(6)`  |
| `r week(7)` | 7     | `r practical(7)`   | `r lecture(7)`  |
| `r week(8)` | 8     | `r practical(8)`   | `r lecture(8)`  |


Exam will take place in the week after the last lecture, please check your timetable for the exact date and location.


# RMD files

You can download the R-Markdown (.Rmd) files for the practical sessions and change and process them in RStudio.
```{r}
#| output: asis
#| echo: false
cat("<ul>")
for (i in 1:visible_practicals)
{
  cat(paste0("<li><a href=\"rmd/data_science_week", i, ".Rmd\">Week ", i, "</a></li>"))
}
cat("</ul>")
```

# Lecture slides

*-- coming soon --*